---
title: "Vegetation Data Ingestion"
date: "`r Sys.Date()`"
output: html_document
params:
  input_file: "data-raw/vegetation_hydrology_data_entry.xlsx"
  project_id: "PROJECT_ID"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(readxl)
library(lubridate)
library(janitor)
library(here)
```

## Purpose

Read raw vegetation field-form data from Excel into two tidy R data frames:
the per-species observations on the `species` sheet, and the per-plot
metadata (location, date, observer, hydrology readings, structural stage,
wetland indicator region, determination, etc.) on the `plots` sheet.
Both feed into `02_qaqc.Rmd`.

## Load species observations

```{r load-species}
veg_raw <- readxl::read_excel(
  here(params$input_file),
  sheet = "species"
)

glimpse(veg_raw)
```

### Standardise species column names

The field-form templates use compact column names (`spp_common`, `spp_name`,
`pct_cover`) optimised for human data entry. Downstream analysis joins against
reference tables that use canonical ecological names (`scientific_name`,
`cover_pct`), so we rename here once at the boundary and every later Rmd works
from the canonical schema. Species are identified by `scientific_name` (the
Latin binomial), which is the join key for every reference table; `common_name`
is carried through purely as a human-readable label. The field crew's
`invasive` flag is preserved as `invasive_field` to keep it distinct from the
reference-table-derived `is_invasive` assigned in `03_transform.Rmd`.

```{r rename-species}
veg <- veg_raw |>
  janitor::clean_names() |>
  dplyr::rename(
    common_name     = spp_common,
    scientific_name = spp_name,
    cover_pct       = pct_cover,
    invasive_field  = invasive
  )
```

## Load plot metadata

The `plots` sheet has one row per field plot. Columns cover identification,
location/date, hydrology readings (in-situ multiprobe + indicators), and
the plot-level outcome calls (`structural_stage` 1-7, `wi_region`,
`determination`, `dom_criterion`). The Excel template enforces a numeric
1-7 dropdown on `structural_stage` so the stage-name labels (Herb / Shrub
/ Sapling / Pole / Young / Mature / Old Forest) can be attached later in
`03_transform.Rmd` or `04_visualize.Rmd` without baking them into the data.

```{r load-plots}
plots_raw <- readxl::read_excel(
  here(params$input_file),
  sheet = "plots"
)

glimpse(plots_raw)
```

### Normalise plot columns

- Parse `date` as a real Date (README convention: text "YYYY-MM-DD").
- Coerce the TRUE/FALSE text dropdowns to logical so downstream code can
  treat them as booleans (`photos_taken`, `water_present`, `ind_*`).
- Coerce numeric chemistry / depth columns defensively in case the dropdown
  validation was bypassed.

```{r normalise-plots}
ind_cols <- grep("^ind_", names(plots_raw), value = TRUE)

plots <- plots_raw |>
  janitor::clean_names() |>
  mutate(
    date = lubridate::ymd(date),
    across(c(photos_taken, water_present, all_of(ind_cols)),
           ~ as.logical(toupper(trimws(.x)))),
    across(c(water_depth_cm, hydro_ph, hydro_ec, hydro_ppm, hydro_temp),
           as.numeric),
    structural_stage = suppressWarnings(as.integer(structural_stage))
  )
```

## Confirm inventory

```{r inventory}
cat("Species sheet:", nrow(veg), "rows across",
    n_distinct(veg$plot_id), "plots.\n")
cat("Plots sheet: ", nrow(plots), "rows.\n")

setdiff(unique(veg$plot_id), unique(plots$plot_id))  # species rows with no plot
setdiff(unique(plots$plot_id), unique(veg$plot_id))  # plots with no species
```

The two `setdiff` calls above flag plot_id mismatches between the sheets —
either typos in transcription or a plot with no species observations.
`02_qaqc.Rmd` formalises these as flags.

## Save interim outputs

```{r save}
saveRDS(veg,   here("data-raw", paste0(params$project_id, "_veg_ingested.rds")))
saveRDS(plots, here("data-raw", paste0(params$project_id, "_plots_ingested.rds")))

message("Saved ingested data for project ", params$project_id, ":\n",
        "  veg:   ", nrow(veg), " species rows\n",
        "  plots: ", nrow(plots), " plot rows")
```
