---
title: "Vegetation Data Ingestion"
date: "`r Sys.Date()`"
output: html_document
params:
  input_file: "data-raw/vegetation_hydrology_data_entry.xlsx"
  project_id: "PROJECT_ID"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(readxl)
library(here)
```

## Purpose

Read raw vegetation field form data from Excel into a tidy R data frame.
This Rmd is the entry point for all vegetation analysis — output feeds into `02_qaqc.Rmd`.

## Load field data

The data entry workbook generated by `inst/templates/create_excel_template.R`
contains two data sheets: `plots` (one row per field plot with hydrology and
plot-level attributes) and `species` (one row per species observation, joined
back to `plots` by `plot_id`). Vegetation analysis works from the species
sheet; plot-level hydrology is consumed by a separate pipeline.

```{r load}
veg_raw <- readxl::read_excel(
  here(params$input_file),
  sheet = "species"
)

glimpse(veg_raw)
```

## Standardise column names

The field-form templates use compact column names (`spp_code`, `spp_name`,
`pct_cover`) optimised for human data entry. Downstream analysis joins against
reference tables that use canonical ecological names (`species_code`,
`scientific_name`, `cover_pct`), so we rename here once at the boundary and
every later Rmd works from the canonical schema. The field crew's `invasive`
flag is preserved as `invasive_field` to keep it distinct from the reference-
table-derived `is_invasive` assigned in `03_transform.Rmd`.

```{r rename}
veg <- veg_raw |>
  janitor::clean_names() |>
  dplyr::rename(
    species_code    = spp_code,
    scientific_name = spp_name,
    cover_pct       = pct_cover,
    invasive_field  = invasive
  )
```

## Confirm plot inventory

```{r plot-summary}
veg |>
  distinct(plot_id) |>
  arrange(plot_id)
```

## Save interim output

```{r save}
saveRDS(veg, here("data-raw", paste0(params$project_id, "_veg_ingested.rds")))
message("Saved: ", nrow(veg), " rows across ", n_distinct(veg$plot_id), " plots.")
```
