---
title: "Vegetation Data QA/QC"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  region: "Central Alberta"  # used to filter region-appropriate species expectations
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(stringr)
library(here)
```

## Load ingested data and reference tables

```{r load}
veg   <- readRDS(here("data-raw", paste0(params$project_id, "_veg_ingested.rds")))
plots <- readRDS(here("data-raw", paste0(params$project_id, "_plots_ingested.rds")))

load(here("data", "wetland_indicator_status.rda"))
load(here("data", "awcs_wetland_species.rda"))
load(here("data", "invasive_species.rda"))
load(here("data", "anpc_wetland_species.rda"))
```

## Reduce species names to binomial

Species names are reduced to their **binomial** (genus + specific epithet)
before any reference lookup. The AWCS list stores names *with* taxonomic
authorities (e.g. `"Betula papyrifera Marsh."`), whereas the WI list and the
field forms use the bare binomial (`"Betula papyrifera"`). Without this
reduction every species fails the AWCS match and is dropped. The binomial is
written back to `scientific_name` so the cleaned output and all downstream joins
use one consistent form. Non-vegetation rows (blank / `"-"` names) are
unaffected.

```{r binomial-names}
# Keep the first two whitespace-separated tokens; 1-token / blank / NA names
# pass through unchanged (sub returns the input when the pattern doesn't match).
to_binomial <- function(x) sub("^(\\S+\\s+\\S+).*$", "\\1", trimws(x))

veg <- veg |> mutate(scientific_name = to_binomial(scientific_name))
```

# Species sheet QA

## Flag: species not in reference tables

Species are identified by `scientific_name` (the Latin binomial), which is the
join key for every reference table in `03_transform.Rmd`. Each species is
checked against the two reference tables — `awcs_wetland_species` (AWCS
class/indicator assignments) and `wetland_indicator_status` (the regional WI
status lookup) — on binomial names (reference names are reduced to binomial the
same way). A species missing from either is flagged, and `missing_in` records
*which* table(s) it was absent from, so the field crew can correct the record
(e.g. a misspelt binomial) or extend the reference table before transform runs.
Non-vegetation rows (bare ground / litter, marked by a blank `scientific_name`)
are excluded from this check.

```{r flag-species}
norm <- function(x) toupper(trimws(x))

known_awcs_names <- norm(to_binomial(awcs_wetland_species$scientific_name))
known_wi_names   <- norm(to_binomial(wetland_indicator_status$scientific_name))

veg_flag_species <- veg |>
  filter(!is.na(scientific_name) & !norm(scientific_name) %in% c("", "-")) |>
  mutate(
    name_in_awcs = norm(scientific_name) %in% known_awcs_names,
    name_in_wi   = norm(scientific_name) %in% known_wi_names
  ) |>
  filter(!name_in_awcs | !name_in_wi) |>
  mutate(
    flag = "UNRECOGNISED_SPECIES",
    missing_in = dplyr::case_when(
      !name_in_awcs & !name_in_wi ~ "AWCS + WI list",
      !name_in_awcs               ~ "AWCS list (data/awcs_wetland_species.rda)",
      !name_in_wi                 ~ "WI list (data/wetland_indicator_status.rda)"
    )
  ) |>
  select(-name_in_awcs, -name_in_wi)

veg_flag_species
```

## Flag: cover values outside expected range (0–100%)

```{r flag-cover}
veg_flag_cover <- veg |>
  filter(!between(cover_pct, 0, 100)) |>
  mutate(flag = "COVER_OUT_OF_RANGE")

veg_flag_cover
```

## Flag: missing required fields (species)

Every row needs `plot_id`, `cover_pct`, and `stratum`. It also needs an
*identity*: vegetation rows carry a `scientific_name`, while non-vegetation
cover rows (bare ground / litter) carry only a `common_name` with a blank
`scientific_name`. A row missing both names has no identity and is flagged.

```{r flag-missing-species}
required_species_fields <- c("plot_id", "cover_pct", "stratum")

blank <- function(x) is.na(x) | trimws(x) %in% c("", "-")

veg_flag_missing <- veg |>
  filter(
    if_any(all_of(required_species_fields), is.na) |
    (blank(scientific_name) & blank(common_name))
  ) |>
  mutate(flag = "MISSING_REQUIRED_FIELD")

veg_flag_missing
```

# Plots sheet QA

The plots sheet is plot-level metadata: one row per plot. Bad rows here
are NOT dropped from `_plots_clean.rds` (that would orphan all of the
plot's species observations); they're flagged for review only.

## Flag: missing required fields (plots)

```{r flag-missing-plots}
required_plot_fields <- c("project_id", "plot_id", "date", "observer")

plots_flag_missing <- plots |>
  filter(if_any(all_of(required_plot_fields), is.na)) |>
  mutate(flag = "PLOT_MISSING_REQUIRED_FIELD")

plots_flag_missing
```

## Flag: duplicate plot_id

```{r flag-dup-plot}
plots_flag_dup <- plots |>
  group_by(plot_id) |>
  filter(dplyr::n() > 1) |>
  ungroup() |>
  mutate(flag = "DUPLICATE_PLOT_ID")

plots_flag_dup
```

## Flag: numeric values outside expected range

Bounds match the dataValidation on the Excel template; double-checked here
in case the validation was bypassed (e.g. paste).

```{r flag-numeric-plots}
plots_flag_numeric <- plots |>
  filter(
    (!is.na(water_depth_cm) & water_depth_cm < 0)        |
    (!is.na(hydro_ph)       & !between(hydro_ph,   0, 14)) |
    (!is.na(hydro_ec)       & hydro_ec   < 0)            |
    (!is.na(hydro_ppm)      & hydro_ppm  < 0)            |
    (!is.na(hydro_temp)     & !between(hydro_temp, -10, 40))
  ) |>
  mutate(flag = "PLOT_NUMERIC_OUT_OF_RANGE")

plots_flag_numeric
```

## Flag: categorical values outside the allowed set

```{r flag-categorical-plots}
valid_stages    <- 1:7
valid_regions   <- c("AK", "WMVC", "GP", "NCNE")
valid_det       <- c("Wetland", "Non-wetland", "Inconclusive")
valid_outlet    <- c("None", "Surface", "Subsurface", "Both")
valid_datum     <- c("NAD83", "WGS84")

plots_flag_categorical <- plots |>
  filter(
    (!is.na(structural_stage) & !structural_stage %in% valid_stages) |
    (!is.na(wi_region)        & !wi_region        %in% valid_regions) |
    (!is.na(determination)    & !determination    %in% valid_det) |
    (!is.na(outlet)           & !outlet           %in% valid_outlet) |
    (!is.na(datum)            & !datum            %in% valid_datum)
  ) |>
  mutate(flag = "PLOT_CATEGORICAL_OUT_OF_SET")

plots_flag_categorical
```

# Cross-sheet QA

## Flag: plot_id in species but not in plots (orphan observations)

```{r flag-orphan-species}
plots_in_data <- unique(plots$plot_id)

veg_flag_orphan <- veg |>
  filter(!plot_id %in% plots_in_data) |>
  mutate(flag = "ORPHAN_SPECIES_NO_PLOT_ROW")

veg_flag_orphan
```

## Flag: plot_id in plots but with no species observations

```{r flag-empty-plot}
species_plots <- unique(veg$plot_id)

plots_flag_empty <- plots |>
  filter(!plot_id %in% species_plots) |>
  mutate(flag = "PLOT_HAS_NO_SPECIES")

plots_flag_empty
```

# QA/QC summary

```{r summary}
veg_flags   <- bind_rows(veg_flag_species, veg_flag_cover,
                          veg_flag_missing, veg_flag_orphan)
plots_flags <- bind_rows(plots_flag_missing, plots_flag_dup,
                          plots_flag_numeric, plots_flag_categorical,
                          plots_flag_empty)

cat("Species sheet:\n")
cat("  Total records  :", nrow(veg), "\n")
cat("  Flagged records:", nrow(veg_flags), "\n")
if (nrow(veg_flags))   print(table(veg_flags$flag))

cat("\nPlots sheet:\n")
cat("  Total records  :", nrow(plots), "\n")
cat("  Flagged records:", nrow(plots_flags), "\n")
if (nrow(plots_flags)) print(table(plots_flags$flag))
```

# Save QA/QC outputs

**Species drop policy.** Only *hard* problems are removed from
`_veg_clean.rds`: out-of-range cover, missing required fields, orphan rows, and
species absent from the **WI list** (whose indicator status the determination
needs). Species present in the WI list but absent from the **AWCS** indicator
list are *flagged* (for the Appendix A AWCS-only annotation) but **retained** —
AWCS is a curated indicator-species list, not a full flora, so absence there is
informational, not a reason to discard valid data. All flags (including
AWCS-only) are still written to the flags CSV. Plots: nothing is dropped from
`_plots_clean.rds` — review the flags CSV and re-transcribe as needed.

```{r save}
# Species absent from the WI list (missing_in mentions "WI list", i.e. WI-only
# or missing-from-both) are dropped; AWCS-only absences are kept.
species_drop <- veg_flag_species |> filter(grepl("WI list", missing_in, fixed = TRUE))
veg_drop     <- bind_rows(species_drop, veg_flag_cover, veg_flag_missing, veg_flag_orphan)

veg_clean   <- veg   |> anti_join(veg_drop, by = names(veg))
plots_clean <- plots                                     # not filtered

message("Species flagged: ", nrow(veg_flags), " | dropped from clean: ",
        nrow(distinct(veg_drop, across(all_of(names(veg))))),
        " | AWCS-only (flagged, kept): ",
        nrow(filter(veg_flag_species, !grepl("WI list", missing_in, fixed = TRUE))))

saveRDS(veg_clean,   here("data-raw", paste0(params$project_id, "_veg_clean.rds")))
saveRDS(plots_clean, here("data-raw", paste0(params$project_id, "_plots_clean.rds")))

write.csv(veg_flags,   here("data-raw", paste0(params$project_id, "_veg_flags.csv")),
          row.names = FALSE)
write.csv(plots_flags, here("data-raw", paste0(params$project_id, "_plots_flags.csv")),
          row.names = FALSE)
```
