---
title: "Vegetation Data Transformation"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  region_priority: "WMVC, GP"
  # Ordered list of regions for wetland indicator status, separated by commas.
  # The first region that has a status for a given species is used; the next
  # region in the list acts as the fallback, and so on. Species with no status
  # in any listed region receive NA and contribute to species richness but no
  # WI-status-derived metric.
  #
  # Valid regions:
  #   WMVC  — Western Mountains & Valleys / Cascades
  #   GP    — Great Plains
  #   NCNE  — North Central & Northeast
  #   AK    — Alaska
  #
  # Guidance for Alberta projects:
  #   Mountains / foothills  → "WMVC, GP"
  #   Parkland / prairies    → "GP, WMVC"
  #   Boreal / Peace region  → "NCNE, GP"
  #   Use all four as safety net → "WMVC, GP, NCNE, AK"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
library(stringr)
library(here)
```

## Purpose

Enrich each species observation with reference-table lookups, then aggregate
to a per-plot summary. Scope per `WORKFLOW.md` §5:

- **Joins**: regional wetland indicator status, native/exotic origin,
  invasive / noxious flags, salinity tolerance.
- **Aggregations**: cover per stratum (T/S/G), bare ground, invasive cover,
  native vs. exotic cover.

Out of scope for this step (deferred): AWIDD dominance test, WAIR scoring,
salinity-composition aggregation. The salinity reference is still joined
onto each row so downstream code (or `04_visualize.Rmd`) can roll it up
when needed.

## Load clean data and reference tables

```{r load}
veg <- readRDS(here("data-raw", paste0(params$project_id, "_veg_clean.rds")))

load(here("data", "wetland_indicator_status.rda"))   # UPL–OBL per region
load(here("data", "anpc_wetland_species.rda"))       # native / exotic origin
load(here("data", "invasive_species.rda"))           # AISC list
load(here("data", "noxious_weeds.rda"))              # AB Weed Control Act
load(here("data", "species_salinity_tolerance.rda")) # ordered most_saline factor
```

## Identify non-vegetation rows

Convention from the field-form template: rows where `scientific_name` is
blank, `NA`, or the literal `"-"` are non-species cover observations
(bare ground, thatch, litter). The `species_code` field tells us which
type (e.g. `"BARE"`, `"THATCH"`, `"LITTER"`); for the per-plot bare-ground
aggregation we collapse all of them.

```{r mark-cover-type}
veg <- veg |>
  mutate(
    sci_clean      = str_trim(scientific_name),
    is_vegetation  = !(is.na(sci_clean) | sci_clean %in% c("", "-")),
    cover_category = if_else(is_vegetation, "vegetation", "bare_or_litter")
  )
```

## Build regional indicator status lookup

```{r region-lookup}
valid_regions <- c("WMVC", "GP", "NCNE", "AK")
region_order  <- trimws(strsplit(params$region_priority, ",")[[1]])

invalid <- setdiff(region_order, valid_regions)
if (length(invalid) > 0)
  stop("Invalid region(s): ", paste(invalid, collapse = ", "),
       ". Must be from: ", paste(valid_regions, collapse = ", "))

message("Region priority: ", paste(region_order, collapse = " -> "))

# wetland_indicator_status.rda columns: scientific_name, common_name,
# wmvc, gp, ncne, ak, species_code. Region columns are lowercase.
region_cols <- tolower(region_order)

wi_lookup <- wetland_indicator_status |>
  select(scientific_name, all_of(region_cols)) |>
  mutate(across(all_of(region_cols), \(x) {
    x <- toupper(trimws(x))
    ifelse(x == "" | x == "NI", NA_character_, x)
  })) |>
  mutate(
    wi_status      = coalesce(!!!syms(region_cols)),
    wi_region_used = {
      m <- pick(all_of(region_cols))
      apply(m, 1, \(row) {
        idx <- which(!is.na(row))[1]
        if (is.na(idx)) NA_character_ else region_order[idx]
      })
    }
  ) |>
  select(scientific_name, wi_status, wi_region_used) |>
  filter(!is.na(wi_status))

message(
  "Indicator status assigned to ", nrow(wi_lookup), " species. ",
  "Coverage by region: ",
  paste(
    sapply(region_order, \(r) paste0(r, "=", sum(wi_lookup$wi_region_used == r, na.rm = TRUE))),
    collapse = ", "
  )
)
```

## Join reference tables onto each species row

```{r join-references}
veg <- veg |>
  # Regional wetland indicator status (UPL–OBL)
  left_join(wi_lookup, by = "scientific_name") |>

  # Native / exotic origin (ANPC)
  left_join(
    anpc_wetland_species |> select(scientific_name, origin),
    by = "scientific_name"
  ) |>

  # Noxious / invasive flags. `invasive_species` is the AISC source list
  # (84 species across prohibited_noxious / noxious / unregulated_invasive
  # categories); `noxious_weeds` holds the richer WCA regulatory metadata
  # for the statutorily regulated subset. The two overlap by design.
  left_join(
    noxious_weeds |> select(scientific_name, weed_class, control_requirement, wca_schedule),
    by = "scientific_name"
  ) |>
  left_join(
    invasive_species |> select(scientific_name, aisc_category),
    by = "scientific_name"
  ) |>
  mutate(
    is_noxious       = !is.na(weed_class),
    is_invasive_aisc = !is.na(aisc_category),
    is_invasive      = is_noxious | is_invasive_aisc
  ) |>

  # Salinity tolerance (joined onto each row; not aggregated here)
  left_join(
    species_salinity_tolerance |>
      select(scientific_name = species_latin, salinity_range, most_saline,
             salinity_source_ids = primary_source_ids),
    by = "scientific_name"
  )
```

## Aggregate per plot

```{r aggregate}
plot_summary <- veg |>
  group_by(plot_id) |>
  summarise(
    # Total vegetation cover per stratum
    cover_tree    = sum(cover_pct[is_vegetation & stratum == "T"], na.rm = TRUE),
    cover_shrub   = sum(cover_pct[is_vegetation & stratum == "S"], na.rm = TRUE),
    cover_ground  = sum(cover_pct[is_vegetation & stratum == "G"], na.rm = TRUE),
    cover_total   = sum(cover_pct[is_vegetation],                  na.rm = TRUE),

    # Non-vegetation cover (bare ground / thatch / litter collapsed)
    cover_bare    = sum(cover_pct[!is_vegetation],                 na.rm = TRUE),

    # Invasive cover (AISC list OR AB Weed Control Act listed)
    cover_invasive = sum(cover_pct[is_vegetation & is_invasive],   na.rm = TRUE),

    # Native vs. exotic cover (per ANPC origin)
    cover_native  = sum(cover_pct[is_vegetation & origin == "Native"], na.rm = TRUE),
    cover_exotic  = sum(cover_pct[is_vegetation & origin == "Exotic"], na.rm = TRUE),
    cover_unknown_origin = sum(cover_pct[is_vegetation &
                                         (is.na(origin) | !origin %in% c("Native","Exotic"))],
                               na.rm = TRUE),

    species_richness = n_distinct(scientific_name[is_vegetation]),

    .groups = "drop"
  )

plot_summary
```

## Save outputs

Two artefacts:

- `_veg_transformed.rds` — every species observation with reference joins
  attached (one row per original observation).
- `_plot_summary.rds` — per-plot aggregations (one row per plot).

```{r save}
saveRDS(veg,          here("data-raw", paste0(params$project_id, "_veg_transformed.rds")))
saveRDS(plot_summary, here("data-raw", paste0(params$project_id, "_plot_summary.rds")))

message("Saved transformed data for ", n_distinct(veg$plot_id), " plots, ",
        nrow(veg), " observations.")
```
