---
title: "Vegetation Data Transformation & Classification"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  region: "WMVC"
  # Region for wetland indicator status lookup.
  # Choices: "WMVC" (Western Mountains & Valleys – Cascades),
  #          "GP"   (Great Plains),
  #          "NCNE" (North Central & Northeast),
  #          "AK"   (Alaska)
  # Most Alberta projects use WMVC (mountains/foothills) or GP (parkland/prairies).
  # When uncertain, use WMVC — it is the most conservative (fewest FAC+ assignments).
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
library(here)
```

## Load clean data and reference tables

```{r load}
veg <- readRDS(here("data-raw", paste0(params$project_id, "_veg_clean.rds")))

load(here("data", "awcs_wetland_species.rda"))
load(here("data", "anpc_wetland_species.rda"))
load(here("data", "invasive_species.rda"))
load(here("data", "noxious_weeds.rda"))
load(here("data", "wair_rules.rda"))
load(here("data", "ecosites_nfc.rda"))

# Alberta Wetland Plant List 2021 — regional wetland indicator status
# Row 1 is a merged title; skip it so row 2 becomes the header.
# Columns: "Scientific Name (ACIMS)", "Common Name (ACIMS)", WMVC, GP, NCNE, AK
ab_plant_list <- read.csv(
  here("data-raw", "ABWetlandPlantList_2021.csv"),
  skip        = 1,
  check.names = FALSE,
  stringsAsFactors = FALSE
)
```

## Select regional indicator status column

```{r region}
region_col <- params$region

stopifnot(
  "`region` must be one of: WMVC, GP, NCNE, AK" =
    region_col %in% c("WMVC", "GP", "NCNE", "AK")
)

message("Using regional indicator column: ", region_col)

# Build a lookup: scientific name → wi_status for the selected region
wi_lookup <- ab_plant_list |>
  select(scientific_name = `Scientific Name (ACIMS)`,
         wi_status        = all_of(region_col)) |>
  filter(!is.na(wi_status), wi_status != "NI") |>
  mutate(wi_status = toupper(trimws(wi_status)))
```

## Join wetland indicator status to vegetation data

```{r join-wi-status}
veg <- veg |>
  left_join(wi_lookup, by = "scientific_name")
```

## Flag noxious weeds and invasive species

```{r join-flags}
veg <- veg |>
  left_join(
    noxious_weeds |> select(scientific_name, weed_class, control_requirement, wca_schedule),
    by = "scientific_name"
  ) |>
  left_join(
    invasive_species |> select(scientific_name, is_invasive = invasive),
    by = "scientific_name"
  ) |>
  mutate(
    is_noxious  = !is.na(weed_class),
    is_invasive = replace_na(is_invasive, FALSE)
  )
```

## Join AWCS wetland class assignments

```{r join-awcs}
veg <- veg |>
  left_join(
    awcs_wetland_species |> select(
      scientific_name,
      awcs_indicator,
      indicator_strength,
      class_bog, class_fen, class_swamp, class_marsh, class_shallow_open_water
    ),
    by = "scientific_name"
  )
```

## WIDD Dominance Test (per plot)

The Alberta Wetland Identification and Delineation Directive defines hydrophytic
vegetation as follows: sort species by cover descending, accumulate until ≥ 50 % of
total plot cover is reached — those are the dominants. If ≥ 50 % of dominant species
have an indicator status of OBL, FACW, or FAC, the vegetation criterion is met.

```{r widd-dominance}
hydrophytic_statuses <- c("OBL", "FACW", "FAC")

plot_wi <- veg |>
  group_by(plot_id) |>
  arrange(desc(cover_pct), .by_group = TRUE) |>
  mutate(
    cumulative_cover = cumsum(cover_pct),
    total_cover      = sum(cover_pct, na.rm = TRUE),
    # A species is dominant if it falls within the 50 % accumulation threshold.
    # Include the species that pushes cumulative cover past 50 % (≥, not >).
    is_dominant = cumulative_cover - cover_pct < total_cover * 0.50
  ) |>
  summarise(
    total_cover           = first(total_cover),
    n_dominant            = sum(is_dominant, na.rm = TRUE),
    n_dominant_hydrophyte = sum(is_dominant & wi_status %in% hydrophytic_statuses,
                                na.rm = TRUE),
    dominant_species      = paste(scientific_name[is_dominant], collapse = "; "),
    .groups = "drop"
  ) |>
  mutate(
    pct_dominant_hydrophyte = n_dominant_hydrophyte / n_dominant,
    vegetation_criterion    = pct_dominant_hydrophyte >= 0.50,
    region_used             = region_col
  )

plot_wi
```

## Summary of weed detections per plot

```{r weed-summary}
weed_detections <- veg |>
  filter(is_noxious | is_invasive) |>
  select(plot_id, scientific_name, common_name, cover_pct,
         weed_class, control_requirement, is_invasive) |>
  arrange(plot_id, weed_class, scientific_name)

weed_detections
```

## Classify plots into ANPC wetland class (WAIR rules)

```{r classify}
# TODO: implement WAIR decision tree using wair_rules reference table
plot_classification <- plot_wi |>
  mutate(
    wetland_class = NA_character_  # to be populated from wair_rules
  )
```

## Save transformed data

```{r save}
saveRDS(veg,                 here("data-raw", paste0(params$project_id, "_veg_transformed.rds")))
saveRDS(plot_wi,             here("data-raw", paste0(params$project_id, "_plot_wi.rds")))
saveRDS(plot_classification, here("data-raw", paste0(params$project_id, "_plot_class.rds")))
saveRDS(weed_detections,     here("data-raw", paste0(params$project_id, "_weed_detections.rds")))
```
