---
title: "Vegetation Data Transformation"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  region_secondary: "GP"
  # Single project-level fallback region for wetland indicator (WI) status.
  # The PRIMARY region is the one recorded at each plot (`plots$wi_region`);
  # this secondary region is used only where the plot's own region returns no
  # status for a species. Leave blank ("") for single-region projects.
  #
  # Valid regions (USACE NWPL regional lists carried in the AB Wetland Plant List):
  #   WMVC  — Western Mountains, Valleys & Coast
  #   GP    — Great Plains
  #   NCNE  — North Central & Northeast
  #   AK    — Alaska
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
library(stringr)
library(here)
```

## Purpose

Enrich each species observation with reference-table lookups and the fields the
hydrophytic determination (`04_analysis.Rmd`) needs, then aggregate to a
per-plot summary. Implements DESIGN §3.

- **Per-plot WI lookup** — wetland indicator status is read from each plot's
  own `wi_region` (primary); where that region has no status for a species, the
  project-level `params$region_secondary` is used as a fallback. The region the
  status was actually drawn from is recorded as `wi_region_used` (drives the
  per-plot table footnote, DESIGN §7).
- **Other joins** — native/exotic origin, invasive / noxious flags, salinity.
- **Determination-support fields** — `stratum_cover`, `is_dominant` (the 20%
  rule), and `wi_numeric` (the prevalence-index weight).
- **Aggregations** — cover per stratum, bare ground, invasive, native vs exotic,
  richness, carried to a per-plot summary.

The wetland determination itself, the % hydrophytic metric, per-wetland rollups,
and the compliance band are computed downstream in `04_analysis.Rmd`.

## Load clean data and reference tables

```{r load}
veg   <- readRDS(here("data-raw", paste0(params$project_id, "_veg_clean.rds")))
plots <- readRDS(here("data-raw", paste0(params$project_id, "_plots_clean.rds")))

load(here("data", "wetland_indicator_status.rda"))   # UPL–OBL per region
load(here("data", "anpc_wetland_species.rda"))       # native / exotic origin
load(here("data", "invasive_species.rda"))           # AISC list
load(here("data", "noxious_weeds.rda"))              # AB Weed Control Act
load(here("data", "species_salinity_tolerance.rda")) # ordered most_saline factor
```

```{r check-wetland-id}
# wetland_id groups plots for every per-wetland figure / rollup (DESIGN §3, §8).
# It is captured explicitly on the plots sheet — never parsed from the plot-ID
# prefix. If the column is absent (older template), warn and carry NA so the
# pipeline still runs; per-wetland rollups will then collapse to a single group.
if (!"wetland_id" %in% names(plots)) {
  warning("`wetland_id` not found on the plots sheet. Per-wetland rollups will ",
          "collapse to NA. Add a `wetland_id` column in the Excel template and ",
          "carry it through 01_ingest.Rmd (DESIGN §3).")
  plots$wetland_id <- NA_character_
}
```

## Identify non-vegetation rows

Convention from the field-form template: rows where `scientific_name` is
blank, `NA`, or the literal `"-"` are non-species cover observations
(bare ground, thatch, litter). The `common_name` field labels which type
(e.g. `"Bare ground"`, `"Thatch"`, `"Litter"`); for the per-plot bare-ground
aggregation we collapse all of them.

```{r mark-cover-type}
veg <- veg |>
  mutate(
    sci_clean      = str_trim(scientific_name),
    is_vegetation  = !(is.na(sci_clean) | sci_clean %in% c("", "-")),
    cover_category = if_else(is_vegetation, "vegetation", "bare_or_litter")
  )
```

## Per-plot wetland indicator status lookup

The reference table `wetland_indicator_status` has one row per species with a
status column per region (`wmvc`, `gp`, `ncne`, `ak`). We reshape it to long
form `(scientific_name, region, wi_status)` and clean the codes — uppercasing
and treating blanks / `"NI"` (no indicator) as missing.

```{r wi-long}
valid_regions <- c("WMVC", "GP", "NCNE", "AK")

region_secondary <- {
  rs <- params$region_secondary
  if (is.null(rs)) "" else toupper(trimws(rs))
}
if (nzchar(region_secondary) && !region_secondary %in% valid_regions)
  stop("Invalid `region_secondary`: ", region_secondary,
       ". Must be one of: ", paste(valid_regions, collapse = ", "), " (or blank).")

wi_long <- wetland_indicator_status |>
  select(scientific_name, wmvc, gp, ncne, ak) |>
  pivot_longer(c(wmvc, gp, ncne, ak), names_to = "region", values_to = "wi_status") |>
  mutate(
    region    = toupper(region),
    wi_status = toupper(trimws(wi_status)),
    wi_status = if_else(wi_status %in% c("", "NI", "NA"), NA_character_, wi_status)
  ) |>
  filter(!is.na(wi_status))

message("Region secondary (fallback): ",
        if (nzchar(region_secondary)) region_secondary else "(none)")
```

For each species row we look up the status in the plot's own region (primary);
where the primary region has no status, we fall back to `region_secondary`, and
record which region the assigned status came from in `wi_region_used`.

```{r wi-lookup}
# Primary: the plot's recorded WI region.
veg <- veg |>
  left_join(
    plots |> transmute(plot_id, plot_wi_region = toupper(trimws(wi_region))),
    by = "plot_id"
  ) |>
  left_join(
    wi_long |> rename(wi_primary = wi_status),
    by = c("scientific_name", "plot_wi_region" = "region")
  )

# Secondary: the single project-level fallback region (may be empty).
wi_secondary_tbl <- wi_long |>
  filter(region == region_secondary) |>
  select(scientific_name, wi_secondary = wi_status)

veg <- veg |>
  left_join(wi_secondary_tbl, by = "scientific_name") |>
  mutate(
    wi_status      = coalesce(wi_primary, wi_secondary),
    wi_region_used = case_when(
      !is.na(wi_primary)   ~ plot_wi_region,
      !is.na(wi_secondary) ~ region_secondary,
      TRUE                 ~ NA_character_
    ),
    # Prevalence-index weight (USACE / Wakeley & Lichvar 1997): OBL=1 … UPL=5.
    wi_numeric = unname(c(OBL = 1, FACW = 2, FAC = 3, FACU = 4, UPL = 5)[wi_status])
  )

message(
  "WI status assigned to ", sum(!is.na(veg$wi_status) & veg$is_vegetation),
  " of ", sum(veg$is_vegetation), " vegetation rows. From secondary region (",
  if (nzchar(region_secondary)) region_secondary else "none", "): ",
  sum(veg$wi_region_used == region_secondary, na.rm = TRUE), " rows."
)
```

## Join remaining reference tables onto each species row

```{r join-references}
veg <- veg |>
  # Native / exotic origin (ANPC)
  left_join(
    anpc_wetland_species |> select(scientific_name, origin),
    by = "scientific_name"
  ) |>

  # Noxious / invasive flags. `invasive_species` is the AISC source list;
  # `noxious_weeds` holds the WCA regulatory metadata for the regulated subset.
  left_join(
    noxious_weeds |> select(scientific_name, weed_class, control_requirement, wca_schedule),
    by = "scientific_name"
  ) |>
  left_join(
    invasive_species |> select(scientific_name, aisc_category),
    by = "scientific_name"
  ) |>
  mutate(
    is_noxious       = !is.na(weed_class),
    is_invasive_aisc = !is.na(aisc_category),
    is_invasive      = is_noxious | is_invasive_aisc
  ) |>

  # Salinity tolerance (joined onto each row; not aggregated here)
  left_join(
    species_salinity_tolerance |>
      select(scientific_name = species_latin, salinity_range, most_saline,
             salinity_source_ids = primary_source_ids),
    by = "scientific_name"
  )
```

## Dominance fields (the 20% rule)

A **dominant** species occupies 20% or more of the total cover of its stratum
within a plot (`methods_vegetation.Rmd`; USDA 2007). `stratum_cover` is the
summed vegetation cover of the stratum in that plot, and `is_dominant` flags
the species meeting the 20% threshold. The 50/20 dominant selection used when a
plot has *no* 20%-rule dominants is computed in `04_analysis.Rmd`.

```{r dominance}
veg <- veg |>
  group_by(plot_id, stratum) |>
  mutate(stratum_cover = sum(cover_pct[is_vegetation], na.rm = TRUE)) |>
  ungroup() |>
  mutate(
    is_dominant = is_vegetation & stratum_cover > 0 &
                  cover_pct >= 0.20 * stratum_cover
  )
```

## Aggregate per plot

```{r aggregate}
plot_summary <- veg |>
  group_by(plot_id) |>
  summarise(
    # Total vegetation cover per stratum
    cover_tree    = sum(cover_pct[is_vegetation & stratum == "T"], na.rm = TRUE),
    cover_shrub   = sum(cover_pct[is_vegetation & stratum == "S"], na.rm = TRUE),
    cover_ground  = sum(cover_pct[is_vegetation & stratum == "G"], na.rm = TRUE),
    cover_total   = sum(cover_pct[is_vegetation],                  na.rm = TRUE),

    # Non-vegetation cover (bare ground / thatch / litter collapsed)
    cover_bare    = sum(cover_pct[!is_vegetation],                 na.rm = TRUE),

    # Invasive cover (AISC list OR AB Weed Control Act listed)
    cover_invasive = sum(cover_pct[is_vegetation & is_invasive],   na.rm = TRUE),

    # Native vs. exotic cover (per ANPC origin)
    cover_native  = sum(cover_pct[is_vegetation & origin == "Native"], na.rm = TRUE),
    cover_exotic  = sum(cover_pct[is_vegetation & origin == "Exotic"], na.rm = TRUE),
    cover_unknown_origin = sum(cover_pct[is_vegetation &
                                         (is.na(origin) | !origin %in% c("Native","Exotic"))],
                               na.rm = TRUE),

    species_richness = n_distinct(scientific_name[is_vegetation]),

    .groups = "drop"
  )

# Attach plot-level attributes from the plots sheet. The field crew's own
# determination / dominance call is carried as `*_field` so it does not clash
# with the determination computed in 04_analysis (and can be QA-compared there).
plot_summary <- plot_summary |>
  left_join(
    plots |> select(plot_id, wetland_id, date, observer,
                    awcs_class_obs, structural_stage, wi_region,
                    determination_field = determination,
                    dom_criterion_field = dom_criterion),
    by = "plot_id"
  )

plot_summary
```

## Save outputs

- `_veg_transformed.rds` — every species observation with reference joins and
  the determination-support fields (`wi_status`, `wi_numeric`, `is_dominant`,
  `stratum_cover`); one row per original observation.
- `_plot_summary.rds` — per-plot aggregations (one row per plot), including
  `wetland_id` and the field-recorded determination for QA.

```{r save}
saveRDS(veg,          here("data-raw", paste0(params$project_id, "_veg_transformed.rds")))
saveRDS(plot_summary, here("data-raw", paste0(params$project_id, "_plot_summary.rds")))

message("Saved transformed data for ", n_distinct(veg$plot_id), " plots, ",
        nrow(veg), " observations.")
```
