---
title: "Vegetation Data Transformation & Classification"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  region_priority: "WMVC, GP"
  # Ordered list of regions for wetland indicator status, separated by commas.
  # The first region that has a status for a given species is used; the next
  # region in the list acts as the fallback, and so on. Species with no status
  # in any listed region receive NA and are excluded from the dominance test
  # without crashing the script.
  #
  # Valid regions:
  #   WMVC  — Western Mountains & Valleys / Cascades
  #   GP    — Great Plains
  #   NCNE  — North Central & Northeast
  #   AK    — Alaska
  #
  # Guidance for Alberta projects:
  #   Mountains / foothills  → "WMVC, GP"
  #   Parkland / prairies    → "GP, WMVC"
  #   Boreal / Peace region  → "NCNE, GP"
  #   Use all four as safety net → "WMVC, GP, NCNE, AK"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
library(here)
```

## Load clean data and reference tables

```{r load}
veg <- readRDS(here("data-raw", paste0(params$project_id, "_veg_clean.rds")))

load(here("data", "awcs_wetland_species.rda"))
load(here("data", "anpc_wetland_species.rda"))
load(here("data", "invasive_species.rda"))
load(here("data", "noxious_weeds.rda"))
load(here("data", "wair_rules.rda"))
load(here("data", "ecosites_nfc.rda"))

# Alberta Wetland Plant List 2021 — regional wetland indicator status
# Row 1 is a merged title; skip it so row 2 becomes the header.
# Columns: "Scientific Name (ACIMS)", "Common Name (ACIMS)", WMVC, GP, NCNE, AK
ab_plant_list <- read.csv(
  here("data-raw", "ABWetlandPlantList_2021.csv"),
  skip        = 1,
  check.names = FALSE,
  stringsAsFactors = FALSE
)
```

## Build regional indicator status lookup (priority order)

```{r region}
valid_regions <- c("WMVC", "GP", "NCNE", "AK")
region_order  <- trimws(strsplit(params$region_priority, ",")[[1]])

invalid <- setdiff(region_order, valid_regions)
if (length(invalid) > 0)
  stop("Invalid region(s): ", paste(invalid, collapse = ", "),
       ". Must be from: ", paste(valid_regions, collapse = ", "))

message("Region priority: ", paste(region_order, collapse = " -> "))

# For each species, pick the first non-empty status in priority order.
# wi_region_used records which column was actually used so results are auditable.
# Species with no status in any listed region get NA and are excluded from the
# dominance test without causing an error.
wi_lookup <- ab_plant_list |>
  select(
    scientific_name = `Scientific Name (ACIMS)`,
    all_of(region_order)
  ) |>
  mutate(across(all_of(region_order), \(x) {
    x <- toupper(trimws(x))
    ifelse(x == "" | x == "NI", NA_character_, x)
  })) |>
  mutate(
    wi_status = coalesce(!!!syms(region_order)),
    wi_region_used = {
      m <- pick(all_of(region_order))
      apply(m, 1, \(row) {
        idx <- which(!is.na(row))[1]
        if (is.na(idx)) NA_character_ else region_order[idx]
      })
    }
  ) |>
  select(scientific_name, wi_status, wi_region_used) |>
  filter(!is.na(wi_status))

message(
  "Indicator status assigned to ", nrow(wi_lookup), " species. ",
  "Coverage by region: ",
  paste(
    sapply(region_order, \(r) paste0(r, "=", sum(wi_lookup$wi_region_used == r, na.rm = TRUE))),
    collapse = ", "
  )
)
```

## Join wetland indicator status to vegetation data

```{r join-wi-status}
# wi_region_used column shows which region's status was applied per species
veg <- veg |>
  left_join(wi_lookup, by = "scientific_name")
```

## Flag noxious weeds and invasive species

```{r join-flags}
veg <- veg |>
  left_join(
    noxious_weeds |> select(scientific_name, weed_class, control_requirement, wca_schedule),
    by = "scientific_name"
  ) |>
  left_join(
    invasive_species |> select(scientific_name, is_invasive = invasive),
    by = "scientific_name"
  ) |>
  mutate(
    is_noxious  = !is.na(weed_class),
    is_invasive = replace_na(is_invasive, FALSE)
  )
```

## Join AWCS wetland class assignments

```{r join-awcs}
veg <- veg |>
  left_join(
    awcs_wetland_species |> select(
      scientific_name,
      common_name,
      awcs_indicator,
      indicator_strength,
      class_bog, class_fen, class_swamp, class_marsh, class_shallow_open_water
    ),
    by = "scientific_name"
  )
```

## WIDD Dominance Test (per plot)

The Alberta Wetland Identification and Delineation Directive defines hydrophytic
vegetation as follows: sort species by cover descending, accumulate until ≥ 50 % of
total plot cover is reached — those are the dominants. If ≥ 50 % of dominant species
have an indicator status of OBL, FACW, or FAC, the vegetation criterion is met.

```{r widd-dominance}
hydrophytic_statuses <- c("OBL", "FACW", "FAC")

plot_wi <- veg |>
  group_by(plot_id) |>
  arrange(desc(cover_pct), .by_group = TRUE) |>
  mutate(
    cumulative_cover = cumsum(cover_pct),
    total_cover      = sum(cover_pct, na.rm = TRUE),
    # A species is dominant if it falls within the 50 % accumulation threshold.
    # Include the species that pushes cumulative cover past 50 % (≥, not >).
    is_dominant = cumulative_cover - cover_pct < total_cover * 0.50
  ) |>
  summarise(
    total_cover           = first(total_cover),
    n_dominant            = sum(is_dominant, na.rm = TRUE),
    n_dominant_hydrophyte = sum(is_dominant & wi_status %in% hydrophytic_statuses,
                                na.rm = TRUE),
    dominant_species      = paste(scientific_name[is_dominant], collapse = "; "),
    .groups = "drop"
  ) |>
  mutate(
    pct_dominant_hydrophyte = n_dominant_hydrophyte / n_dominant,
    vegetation_criterion    = pct_dominant_hydrophyte >= 0.50,
    region_priority         = paste(region_order, collapse = " > ")
  )

plot_wi
```

## Summary of weed detections per plot

```{r weed-summary}
weed_detections <- veg |>
  filter(is_noxious | is_invasive) |>
  select(plot_id, scientific_name, common_name, cover_pct,
         weed_class, control_requirement, is_invasive) |>
  arrange(plot_id, weed_class, scientific_name)

weed_detections
```

## Classify plots into ANPC wetland class (WAIR rules)

```{r classify}
# TODO: implement WAIR decision tree using wair_rules reference table
plot_classification <- plot_wi |>
  mutate(
    wetland_class = NA_character_  # to be populated from wair_rules
  )
```

## Save transformed data

```{r save}
saveRDS(veg,                 here("data-raw", paste0(params$project_id, "_veg_transformed.rds")))
saveRDS(plot_wi,             here("data-raw", paste0(params$project_id, "_plot_wi.rds")))
saveRDS(plot_classification, here("data-raw", paste0(params$project_id, "_plot_class.rds")))
saveRDS(weed_detections,     here("data-raw", paste0(params$project_id, "_weed_detections.rds")))
```
