---
title: "Vegetation Analysis — Hydrophytic Determination"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  compliance_band_pct: 10
  # Compliance band half-width (DESIGN §4). Applied as percentage POINTS for
  # cover metrics and as a percentage of the baseline count (rounded up) for
  # species richness. Computed here for the current survey instance (= baseline);
  # WAIR stores it unused, Monitoring tests later years against it.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
library(tibble)
library(here)
```

## Purpose

The regulatory core (DESIGN §3). From the per-species table and per-plot summary
produced by `03_transform.Rmd`, this step:

1. runs the **tiered three-method hydrophytic determination** per plot
   (`methods_vegetation.Rmd`; USDA 2007, Wakeley & Lichvar 1997, USACE 2018);
2. computes the **% hydrophytic vegetation** headline metric;
3. rolls per-plot metrics up to **per-wetland** means (± SD); and
4. computes the **baseline compliance band** for every plot × metric and
   wetland × metric (computed here, drawn only by Monitoring).

It produces the determination outputs every figure and table downstream depends
on. It does **not** draw anything.

```{r load}
veg          <- readRDS(here("data-raw", paste0(params$project_id, "_veg_transformed.rds")))
plot_summary <- readRDS(here("data-raw", paste0(params$project_id, "_plot_summary.rds")))

band_pct <- as.numeric(params$compliance_band_pct)
hydrophytic_status <- c("OBL", "FACW", "FAC")
```

## Hydrophytic determination — the tiered three-method test

A **dominant** is a species ≥ 20% of its stratum cover (`is_dominant`, from
`03_transform`). The tier is:

1. **Method 1 — dominance of identified dominants.** If a plot has any 20%-rule
   dominants, classify on them: > 50% hydrophytic (OBL/FACW/FAC) ⇒ wetland,
   < 50% ⇒ upland. An **exact 50/50 split is a tie** and cascades to Method 2.
2. **Method 2 — 50/20 rule (USDA 2007).** Applied when a plot has *no* 20%-rule
   dominants **or** Method 1 tied. Select dominants by the 50/20 rule (rank by
   cover within stratum; take species cumulatively to 50% of the stratum total,
   plus any other species exceeding 20%), then apply the dominance test: > 50%
   ⇒ wetland, < 50% ⇒ upland; an exact 50/50 split cascades to Method 3.
3. **Method 3 — prevalence index (Wakeley & Lichvar 1997).**
   `PI = Σ(cover·weight) / Σ(cover)` over rated species; `PI < 3` ⇒ wetland,
   otherwise upland. The final fail-safe — it cannot tie.

> **Cascade of fail-safes.** The three methods form a cascade: each is decisive
> unless its dominance test lands on an exact 50/50 tie, in which case the next
> method is applied. Method 1 (20%-rule dominants) → on a tie → Method 2 (50/20
> dominants) → on a tie → Method 3 (prevalence index). The method that actually
> resolves the plot is recorded in `deciding_method`.

```{r dom-5020}
# 50/20 dominant selection within one stratum's cover vector.
dom_5020 <- function(cover) {
  cover[is.na(cover)] <- 0
  total <- sum(cover)
  if (total <= 0) return(rep(FALSE, length(cover)))
  ord <- order(cover, decreasing = TRUE)
  cum <- cumsum(cover[ord])
  n50 <- which(cum >= 0.5 * total)[1]   # species cumulatively reaching 50%
  sel <- logical(length(cover))
  sel[ord[seq_len(n50)]] <- TRUE
  sel[cover > 0.2 * total] <- TRUE      # plus any species exceeding 20%
  sel
}

veg_v <- veg |>
  filter(is_vegetation) |>
  group_by(plot_id, stratum) |>
  mutate(is_dom_5020 = dom_5020(cover_pct)) |>
  ungroup()
```

```{r determination}
det_raw <- veg_v |>
  group_by(plot_id) |>
  summarise(
    # Method 1 dominant set (20% rule)
    n_dom_m1 = sum(is_dominant, na.rm = TRUE),
    n_hyd_m1 = sum(is_dominant & wi_status %in% hydrophytic_status, na.rm = TRUE),
    # Method 2 dominant set (50/20 rule)
    n_dom_m2 = sum(is_dom_5020, na.rm = TRUE),
    n_hyd_m2 = sum(is_dom_5020 & wi_status %in% hydrophytic_status, na.rm = TRUE),
    # Prevalence index over rated species (all strata)
    pi_num   = sum(cover_pct * wi_numeric, na.rm = TRUE),
    pi_den   = sum(cover_pct[!is.na(wi_numeric)], na.rm = TRUE),
    .groups  = "drop"
  )

# Anchor on the full plot list so plots with NO vegetation rows (e.g. bare
# ground only, or all species dropped in QA) still get a row and resolve as
# Inconclusive — rather than silently disappearing from the determination.
det <- plot_summary |>
  distinct(plot_id) |>
  left_join(det_raw, by = "plot_id") |>
  mutate(across(c(n_dom_m1, n_hyd_m1, n_dom_m2, n_hyd_m2, pi_num, pi_den),
                ~ coalesce(.x, 0))) |>
  mutate(
    # Dominance-test result for each method's dominant set (NA if that set is
    # empty). pct == 50 is exact for integer dominant counts, so the tie test
    # below is safe.
    pct1 = if_else(n_dom_m1 > 0, 100 * n_hyd_m1 / n_dom_m1, NA_real_),
    pct2 = if_else(n_dom_m2 > 0, 100 * n_hyd_m2 / n_dom_m2, NA_real_),
    prevalence_index = if_else(pi_den > 0, pi_num / pi_den, NA_real_),

    # Cascade of fail-safes: Method 1 decides unless it ties at exactly 50%,
    # then Method 2, then Method 3 (which cannot tie).
    deciding_method = case_when(
      n_dom_m1 == 0 & n_dom_m2 == 0 ~ "Inconclusive: no vascular species",
      n_dom_m1 > 0 & pct1 != 50     ~ "Method 1: dominance test",
      n_dom_m2 > 0 & pct2 != 50     ~ "Method 2: 50/20 dominance",
      TRUE                          ~ "Method 3: prevalence index"
    ),
    determination = case_when(
      deciding_method == "Inconclusive: no vascular species" ~ "Inconclusive",
      deciding_method == "Method 1: dominance test"  ~ if_else(pct1 > 50, "Wetland", "Non-wetland"),
      deciding_method == "Method 2: 50/20 dominance" ~ if_else(pct2 > 50, "Wetland", "Non-wetland"),
      TRUE ~ if_else(!is.na(prevalence_index) & prevalence_index < 3, "Wetland", "Non-wetland")
    ),

    # The dominant set / metrics reported are the ones the deciding method used
    # (Method 3 reports its Method-2 dominant set, which tied at 50%).
    dominant_basis = case_when(
      deciding_method == "Method 1: dominance test" ~ "20% rule",
      startsWith(deciding_method, "Method 2") |
        startsWith(deciding_method, "Method 3")     ~ "50/20 rule",
      TRUE                                          ~ NA_character_
    ),
    n_dominants = case_when(
      deciding_method == "Method 1: dominance test" ~ n_dom_m1,
      startsWith(deciding_method, "Method 2") |
        startsWith(deciding_method, "Method 3")     ~ n_dom_m2,
      TRUE                                          ~ 0L
    ),
    n_dominants_hydrophytic = case_when(
      deciding_method == "Method 1: dominance test" ~ n_hyd_m1,
      startsWith(deciding_method, "Method 2") |
        startsWith(deciding_method, "Method 3")     ~ n_hyd_m2,
      TRUE                                          ~ 0L
    ),
    pct_hydrophytic = case_when(
      deciding_method == "Method 1: dominance test" ~ pct1,
      startsWith(deciding_method, "Method 2") |
        startsWith(deciding_method, "Method 3")     ~ pct2,
      TRUE                                          ~ NA_real_
    )
  )
```

```{r plot-class}
plot_class <- det |>
  left_join(plot_summary |> select(plot_id, wetland_id), by = "plot_id") |>
  select(plot_id, wetland_id, determination, deciding_method, dominant_basis,
         n_dominants, n_dominants_hydrophytic, pct_hydrophytic, prevalence_index)

plot_class
```

### QA: computed vs field-recorded determination

The field crew's own call (`determination_field`) is compared against the
computed determination; mismatches are surfaced for review (not corrected).

```{r qa-determination}
det_compare <- plot_class |>
  left_join(plot_summary |> select(plot_id, determination_field), by = "plot_id") |>
  mutate(agrees = determination == determination_field) |>
  select(plot_id, determination, determination_field, deciding_method, agrees)

mismatches <- det_compare |> filter(!agrees | is.na(agrees))
cat("Plots:", nrow(det_compare),
    "| computed vs field mismatches/NA:", nrow(mismatches), "\n")
mismatches
```

## % hydrophytic and per-plot metrics

`pct_hydrophytic` joins onto the per-plot summary as the headline determination
metric (the `Dominance_Plot` y-axis), alongside the computed determination.

```{r enrich-plot-summary}
plot_summary <- plot_summary |>
  left_join(
    plot_class |> select(plot_id, determination, deciding_method,
                         dominant_basis, pct_hydrophytic),
    by = "plot_id"
  )
```

## Compliance band (computed, not drawn)

Per DESIGN §4. Cover metrics get an absolute ± band in percentage points;
species richness gets ± a percentage of the baseline count, rounded up. The
adverse direction is metric-specific (a *decrease* is adverse for cover/native/
richness; an *increase* is adverse for invasive/bare/exotic). The lower bound is
clamped at 0; cover has no upper clamp (overlapping strata can exceed 100%).

```{r metric-meta}
metric_meta <- tribble(
  ~metric,            ~band_type, ~adverse_direction,
  "cover_total",      "cover",    "decrease",
  "cover_bare",       "cover",    "increase",
  "cover_invasive",   "cover",    "increase",
  "cover_native",     "cover",    "decrease",
  "cover_exotic",     "cover",    "increase",
  "species_richness", "count",    "decrease"
)

band_half_width <- function(value, band_type, band_pct) {
  if_else(band_type == "cover",
          band_pct,                              # percentage points
          ceiling(band_pct / 100 * value))       # % of count, rounded up
}
```

```{r plot-bands}
plot_bands <- plot_summary |>
  select(plot_id, wetland_id, all_of(metric_meta$metric)) |>
  pivot_longer(all_of(metric_meta$metric), names_to = "metric", values_to = "value") |>
  left_join(metric_meta, by = "metric") |>
  mutate(
    half_width = band_half_width(value, band_type, band_pct),
    band_lo    = pmax(0, value - half_width),
    band_hi    = value + half_width
  )

head(plot_bands, 12)
```

## Per-wetland rollups

Mean (± SD) across the plots in each `wetland_id` — the WAIR per-wetland error
bar (descriptive variability). The wetland-level baseline band (mean ± band) is
also carried for Monitoring's per-wetland compliance test.

```{r wetland-summary}
wetland_summary <- plot_summary |>
  select(wetland_id, all_of(metric_meta$metric)) |>
  pivot_longer(all_of(metric_meta$metric), names_to = "metric", values_to = "value") |>
  group_by(wetland_id, metric) |>
  summarise(
    n_plots = sum(!is.na(value)),
    mean    = mean(value, na.rm = TRUE),
    sd      = sd(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(metric_meta, by = "metric") |>
  mutate(
    half_width = band_half_width(mean, band_type, band_pct),
    band_lo    = pmax(0, mean - half_width),
    band_hi    = mean + half_width
  )

wetland_summary
```

## Save outputs

- `_plot_class.rds` — per-plot determination: `determination`, `deciding_method`,
  `dominant_basis`, dominant counts, `pct_hydrophytic`, `prevalence_index`.
- `_plot_summary.rds` — per-plot metrics enriched with the determination and
  `pct_hydrophytic` (overwrites the `03_transform` summary).
- `_plot_bands.rds` — per plot × metric baseline compliance band (long).
- `_wetland_summary.rds` — per wetland × metric mean ± SD and baseline band.

> Note: DESIGN §3 folds the per-wetland rollups conceptually into
> `_plot_summary.rds`. They are saved here as separate `_plot_bands.rds` /
> `_wetland_summary.rds` files (long, one row per metric) so the band model is
> tidy and `06_visualize` / `05_monitoring` can read exactly the level they need.

```{r save}
saveRDS(plot_class,      here("data-raw", paste0(params$project_id, "_plot_class.rds")))
saveRDS(plot_summary,    here("data-raw", paste0(params$project_id, "_plot_summary.rds")))
saveRDS(plot_bands,      here("data-raw", paste0(params$project_id, "_plot_bands.rds")))
saveRDS(wetland_summary, here("data-raw", paste0(params$project_id, "_wetland_summary.rds")))

message("Determination complete for ", nrow(plot_class), " plots: ",
        paste(names(table(plot_class$determination)),
              table(plot_class$determination), sep = "=", collapse = ", "))
```
