<!--
methods_vegetation.Rmd — verbatim vegetation analysis methodology for the
report appendix. Intended to be included as a knitr CHILD document, e.g.:

    ```{r veg-methods, child = here::here("analysis/vegetation/methods_vegetation.Rmd")}
    ```

so it has no YAML front matter and inherits the parent report's `params`.
It also renders standalone for preview (the setup chunk supplies fallbacks).

Every method claim is traceable to a `source_id` in `data/references.rda`
(see "Source references" at the foot). All citations the firm's documented
method relies on are now present in references.csv:
  * usda_2007            — the 50/20 dominance methodology ("USDA 2007")
  * wakeley_lichvar_1997 — the prevalence index ("Wakeley & Lichvar 1997")
The prevalence index previously cited USACE 2011, which is no longer available
online; it is replaced by Wakeley & Lichvar (1997), the source paper for the
plot-based prevalence index, with De Steven (2015) as a clarifying note.
-->

```{r veg-methods-setup, include=FALSE}
# Project-specific values come from the parent report's params; fall back to
# placeholders so the document also renders on its own for preview/QA.
.mp <- function(name, default) {
  if (exists("params") && is.list(params) && !is.null(params[[name]])) {
    params[[name]]
  } else {
    default
  }
}
region_secondary <- .mp("region_secondary", "GP")
band_pct         <- .mp("compliance_band_pct", 10)
report_type      <- .mp("report_type", "wair")
```

## Vegetation

### Field sampling

Plant species were assessed for percent cover within each vegetation stratum at
every vegetation plot, using nested plots sized to the stratum: **1 m × 1 m** at
the ground (herb) stratum, **5 m × 5 m** at the shrub stratum, and
**10 m × 10 m** at the tree stratum. Percent cover of dominant vascular species,
total vascular species, litter, bare ground, and open water was recorded where
applicable. Plot layout and delineation follow the Alberta Wetland
Identification and Delineation Directive (AWIDD; GoA 2015b).

A **dominant vascular species** is defined as a species that occupies **20% or
more** of the areal coverage of its respective stratum.

### Wetland indicator status

To identify hydrophytic plant species, species observed in the field were
referenced against the **Alberta Wetland Plant List (2021)**. Because no
Alberta-specific reference of comparable detail is currently available, it is
common practice to reference the **National Wetland Plant List (NWPL)** when
rating wetland indicator status; the Alberta Wetland Plant List is itself
adapted from the NWPL regional lists.

Wetland indicator status was assigned using the regional plant list for the
**wetland indicator region recorded at each plot** (the primary region). Where
the regional list returned no rating for a species in that plot's region, the
status from the **secondary region (`r region_secondary`)** was used as a
fallback. Any rating drawn from the secondary region is flagged with a
superscript in the per-plot vegetation table (Appendix A).

Species are rated in one of five categories based on their affinity for wetland
versus upland conditions (Table 6). For the purposes of the wetland
determination below, species rated **obligate (OBL)**, **facultative wetland
(FACW)**, or **facultative (FAC)** are treated as hydrophytic.

*Table 6. Wetland Indicator Status Ratings*

| Indicator Status | Ecological Description (USACE 2018) |
|---|---|
| Obligate (OBL) | Almost always is a hydrophyte, rarely found in uplands |
| Facultative Wetland (FACW) | Usually is a hydrophyte but occasionally found in uplands |
| Facultative (FAC) | Commonly occurs as either a hydrophyte or non-hydrophyte |
| Facultative Upland (FACU) | Occasionally is a hydrophyte, but usually occurs in uplands |
| Upland (UPL) | Rarely a hydrophyte, almost always in uplands |

Species with no indicator status under the NWPL but classified under the Alberta
Wetland Classification System (AWCS; GoA 2015a) are noted in Appendix A (e.g.
`M` = species characteristic of marshes).

### Hydrophytic vegetation and wetland determination

To determine whether a plot was located within a wetland, a **tiered approach
involving three methods** was applied. Method 1 was applied first; Methods 2 and
3 were employed sequentially only where the preceding method did not yield a
conclusive result.

1. **Dominance of identified dominants.** If more than 50% of the dominant
   species across all strata are obligate (OBL), facultative wetland (FACW), or
   facultative (FAC), the plot is considered to occur within a wetland.

2. **50/20 dominance methodology (USDA 2007).** If no dominant species were
   identified in the plot, the 50/20 methodology was implemented to determine
   dominance — ranking species by percent cover within each stratum, selecting
   species cumulatively until 50% of the stratum total is reached, plus any
   additional species exceeding 20% cover — and the dominance test re-applied.
   Where this produced a tie on the 50% dominance test, Method 3 was run.

3. **Prevalence index (Wakeley & Lichvar 1997).** The prevalence index (PI)
   was calculated to resolve the tied dominance test of Method 2 (see also
   De Steven 2015):

   $$ PI = \frac{\sum_{i} A_i \, w_i}{\sum_{i} A_i} $$

   where $A_i$ is the percent cover of species $i$ and $w_i$ is its indicator
   weight (OBL = 1, FACW = 2, FAC = 3, FACU = 4, UPL = 5). A plot with
   **PI < 3** is considered to occur within a wetland.

The method that produced each plot's determination is retained so every wetland
call is traceable (Appendix A, per-plot determination summary).

### Derived vegetation metrics

The following metrics were summarised per plot and per wetland:

- **Total, tree, shrub, and ground cover** — summed percent cover of vascular
  vegetation by stratum and overall.
- **Bare ground cover** — combined non-vegetation cover (bare ground, litter,
  thatch).
- **Native versus exotic cover** — percent cover by origin (Alberta Native Plant
  Council; ANPC 2023).
- **Invasive and noxious cover** — percent cover of species listed by the
  Alberta Invasive Species Council (AISC 2022) or regulated under the Alberta
  *Weed Control Act* (GoA 2023).
- **Species richness** — count of distinct vascular species.
- **Percent hydrophytic vegetation** — the proportion of dominant species rated
  OBL, FACW, or FAC, expressed as a percentage; the primary indicator tracked in
  the dominance test.

```{r veg-methods-monitoring, eval = identical(report_type, "monitoring"), echo = FALSE, results = "asis"}
cat(sprintf('
### Compliance assessment (wetland monitoring)

For monitoring projects, each metric was assessed against a compliance band
fixed from the **baseline year**. The band is an absolute **±%d**% of the
baseline value for cover metrics (in percentage points) and
**±⌈%d%% × baseline count⌉** species for richness (rounded up to the
nearest whole species). Both plot-level and wetland-level values were evaluated.

A wetland is considered to be trending toward **non-compliance** where a metric
moves past the band in the adverse direction for that metric: a **decrease**
below the lower bound for vegetation cover, native cover, and species richness;
an **increase** above the upper bound for invasive and bare-ground cover.
Movement in the favourable direction (for example, increased desirable cover) is
not a compliance exceedance.
', band_pct, band_pct))
```

### Source references

Citations resolve to `source_id` entries in `data/references.rda`:

| In-text citation | `source_id` |
|---|---|
| Alberta Wetland Plant List (2021) | `ab_wetland_plant_list_2021` |
| National Wetland Plant List (NWPL) | `usace_nwpl_2018` |
| Indicator rating definitions (USACE 2018) | `usace_nwpl_2018`, `lichvar_2012` |
| AWCS (GoA 2015a) | `goa_awcs_2015` |
| AWIDD (GoA 2015b) | `goa_awidd` |
| ANPC (2023) | `anpc_2023` |
| AISC (2022) | `aisc_2022` |
| *Weed Control Act* (GoA 2023) | `wca_2023` |
| 50/20 methodology (USDA 2007) | `usda_2007` |
| Prevalence index (Wakeley & Lichvar 1997) | `wakeley_lichvar_1997` |
| Prevalence index clarification (De Steven 2015) | `de_steven_2015` |
