---
title: "Peat Data Ingestion"
date: "`r Sys.Date()`"
output: html_document
params:
  input_file: "data-raw/projects/peat_data_entry.xlsx"
  project_id: "PROJECT_ID"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(readxl)
library(janitor)
library(here)
```

The peat data-entry workbook (`inst/templates/peat_data_entry.xlsx`) has two
sheets:

  - **plots**    — one row per test hole (header band of the field sheet)
  - **horizons** — one row per peat interval (the field-sheet table)

Both have a title banner in row 1, column headers in row 2, and a grey example
row in row 3. We read with `skip = 1` so row 2 becomes the header, then drop the
example/blank rows. Delete the grey example rows before submitting; the
`!is.na(plot_id)` filter below is only a backstop for blank trailing rows.

## Load both sheets

```{r load}
plots_raw <- readxl::read_excel(here(params$input_file), sheet = "plots",
                                skip = 1) |>
  janitor::clean_names()

horizons_raw <- readxl::read_excel(here(params$input_file), sheet = "horizons",
                                   skip = 1) |>
  janitor::clean_names()

glimpse(plots_raw)
glimpse(horizons_raw)
```

## Coerce types and drop blank rows

```{r coerce}
plots <- plots_raw |>
  filter(!is.na(plot_id)) |>
  mutate(across(c(gps_easting, gps_northing, elevation_m,
                  water_table_depth_m, ph, conductivity_us_cm),
                as.numeric),
         date = as.character(date))

horizons <- horizons_raw |>
  filter(!is.na(plot_id)) |>
  mutate(across(c(depth_top_cm, depth_bot_cm), as.numeric),
         von_post = toupper(trimws(von_post)))
```

## Confirm test-hole inventory

```{r inventory}
# Every horizon's plot_id should appear in plots, and vice versa.
plots |> distinct(plot_id) |> arrange(plot_id)

cat("Plots:", nrow(plots),
    "| Holes with intervals:", dplyr::n_distinct(horizons$plot_id),
    "| Interval records:", nrow(horizons), "\n")
```

## Save

```{r save}
# Keep the two tables paired in a single list so downstream steps load one file.
peat <- list(plots = plots, horizons = horizons)
saveRDS(peat, here("data-raw", paste0(params$project_id, "_peat_ingested.rds")))
```
