---
title: "Peat Data QA/QC"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(here)
```

## Load data and reference tables

```{r load}
peat     <- readRDS(here("data-raw", paste0(params$project_id, "_peat_ingested.rds")))
plots    <- peat$plots
horizons <- peat$horizons

load(here("data", "von_post.rda"))                      # von_post_code = "H1".."H10"
ecosite_wetclass <- read.csv(
  here("data-raw", "Ecosite_to_WetClass.csv"), check.names = FALSE)
valid_classes <- trimws(ecosite_wetclass[[1]])          # first column = AWCS class/form
valid_classes <- valid_classes[valid_classes != "" &
                                 valid_classes != "AB Wetland Classification Class and Form"]
# "Upland" is a valid non-wetland designation for a test hole sited outside a
# wetland. It is not an AWCS class, so add it explicitly to match the form's
# dropdown (inst/templates/create_peat_template.R).
valid_classes <- c(valid_classes, "Upland")
```

> **Note on non-monotonic cores.** Von Post H values legitimately flip with
> depth (peat stratigraphy is not monotonic). We therefore do **not** flag
> non-monotonic H sequences. We enforce *depth* contiguity only.

## Flag: inverted or non-contiguous interval depths

```{r flag-depths}
# Within each hole, the sub-surface intervals must start at 0 and be contiguous:
#   each depth_top_cm == previous depth_bot_cm, and depth_top < depth_bot.
# Forest-floor (LFH and L/F/H folic) horizons are recorded as thickness ABOVE
# the surface (e.g. 15 -> 0), so they are exempt from the inversion/contiguity
# rules and excluded from the depth sequence below.
depth_checked <- horizons |>
  filter(!grepl("^(LFH|LF|FH|L|F|H)$", toupper(trimws(horizon)))) |>
  arrange(plot_id, depth_top_cm) |>
  group_by(plot_id) |>
  mutate(prev_bot   = lag(depth_bot_cm),
         expected_top = ifelse(is.na(prev_bot), 0, prev_bot)) |>
  ungroup()

flag_inverted <- depth_checked |>
  filter(depth_top_cm >= depth_bot_cm) |>
  mutate(flag = "INVERTED_INTERVAL_DEPTH")

flag_gap <- depth_checked |>
  filter(abs(depth_top_cm - expected_top) > 1e-6) |>
  mutate(flag = "NON_CONTIGUOUS_INTERVAL")

bind_rows(flag_inverted, flag_gap) |> select(plot_id, depth_top_cm, depth_bot_cm, flag)
```

## Flag: missing or invalid Von Post on a peat interval

```{r flag-vonpost}
valid_vp <- von_post |> pull(von_post_code)

# Von Post humification applies only to organic PEAT horizons (Of/Om/Oh, ...).
# Mineral (C/Cg, A/B) and forest-floor (LFH) horizons have no humification value,
# so a blank von_post on them is expected, not an error.
flag_vp_missing <- horizons |>
  filter(grepl("^O", toupper(trimws(horizon))),
         is.na(von_post) | von_post == "") |>
  mutate(flag = "MISSING_VON_POST")

flag_vp_invalid <- horizons |>
  filter(!is.na(von_post), von_post != "", !von_post %in% valid_vp) |>
  mutate(flag = "INVALID_VON_POST")

bind_rows(flag_vp_missing, flag_vp_invalid) |> select(plot_id, depth_top_cm, von_post, flag)
```

## Flag: plot-level out-of-range or unrecognised values

```{r flag-plots}
flag_class <- plots |>
  filter(!is.na(awcs_wetland_class), !awcs_wetland_class %in% valid_classes) |>
  mutate(flag = "UNRECOGNISED_WETLAND_CLASS")

flag_zone <- plots |>
  filter(!utm_zone %in% c("11N", "12N")) |>
  mutate(flag = "INVALID_UTM_ZONE")

flag_ph <- plots |>
  filter(!is.na(ph), ph < 0 | ph > 14) |>
  mutate(flag = "PH_OUT_OF_RANGE")

flag_coords <- plots |>
  filter(is.na(gps_easting) | is.na(gps_northing)) |>
  mutate(flag = "MISSING_COORDINATES")

flag_plotlevel <- bind_rows(flag_class, flag_zone, flag_ph, flag_coords)
flag_plotlevel |> select(plot_id, awcs_wetland_class, utm_zone, ph, flag)
```

## Flag: water table deeper than the peat profile

```{r flag-watertable}
peat_depth <- horizons |>
  group_by(plot_id) |>
  summarise(mineral_depth_cm = max(depth_bot_cm, na.rm = TRUE), .groups = "drop")

flag_wt <- plots |>
  left_join(peat_depth, by = "plot_id") |>
  filter(!is.na(water_table_depth_m),
         water_table_depth_m * 100 > mineral_depth_cm) |>
  mutate(flag = "WATER_TABLE_BELOW_MINERAL")

flag_wt |> select(plot_id, water_table_depth_m, mineral_depth_cm, flag)
```

## Flag: referential integrity between sheets

```{r flag-referential}
holes_in_plots    <- unique(plots$plot_id)
holes_in_horizons <- unique(horizons$plot_id)

flag_orphan_horizon <- horizons |>
  filter(!plot_id %in% holes_in_plots) |>
  mutate(flag = "HORIZON_WITHOUT_PLOT")

# An upland test hole has no peat profile by design, so absence of horizons is
# expected there — don't flag plots classed "Upland".
flag_plot_no_horizon <- plots |>
  filter(!plot_id %in% holes_in_horizons,
         !(awcs_wetland_class %in% "Upland")) |>
  mutate(flag = "PLOT_WITHOUT_HORIZONS")

bind_rows(
  flag_orphan_horizon |> select(plot_id, flag),
  flag_plot_no_horizon |> select(plot_id, flag)
)
```

## Summary and save

```{r save}
horizon_flags <- bind_rows(flag_inverted, flag_gap, flag_vp_missing,
                           flag_vp_invalid, flag_orphan_horizon) |>
  select(plot_id, depth_top_cm, depth_bot_cm, von_post, flag)
plot_flags    <- bind_rows(flag_plotlevel, flag_wt, flag_plot_no_horizon) |>
  select(plot_id, flag)

all_flags <- bind_rows(horizon_flags, plot_flags)
cat("Total flags raised:", nrow(all_flags), "\n")
print(table(all_flags$flag))

# Clean sets = rows not implicated in a structural flag. Plot-level value flags
# (pH range, unrecognised class) are surfaced for review but not auto-dropped,
# so the analyst can decide; structural flags (bad depths, orphans) are removed.
bad_horizons <- bind_rows(flag_inverted, flag_gap, flag_orphan_horizon)
bad_plots    <- bind_rows(flag_plot_no_horizon)

horizons_clean <- horizons |> anti_join(bad_horizons, by = names(horizons))
plots_clean    <- plots    |> anti_join(bad_plots,    by = names(plots))

peat_clean <- list(plots = plots_clean, horizons = horizons_clean)
saveRDS(peat_clean, here("data-raw", paste0(params$project_id, "_peat_clean.rds")))
write.csv(all_flags,
          here("data-raw", paste0(params$project_id, "_peat_flags.csv")),
          row.names = FALSE)
```
