---
title: "Peat Transformation, Volume & Spatial Interpolation"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  approved_boundary: "data-raw/projects/approved_harvest_area.shp"
  ownership_boundary: "data-raw/projects/ownership_boundary.shp"
  reclaim_depth_cm: 50      # material left on the mineral substrate for reclamation
  raster_res_m: 0           # 0 = auto (mean nearest-neighbour core spacing / 10, clamped 5-25 m)
  idw_power: 2              # IDW distance-decay exponent
  idw_radius_m: 0           # 0 = auto (full extent diagonal → global IDW)
  output_dir: "reports/figures"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(sf)
library(terra)
library(here)
dir.create(here(params$output_dir), showWarnings = FALSE, recursive = TRUE)
```

## Load clean data

```{r load}
peat     <- readRDS(here("data-raw", paste0(params$project_id, "_peat_clean.rds")))
plots    <- peat$plots
horizons <- peat$horizons
```

## Band each interval by Von Post humification

Banding is **independent of vertical order**, so non-monotonic cores need no
special handling — each interval is assigned by its own H value.

  - **H1–H4 → professional** (premium horticultural grade)
  - **H5–H6 → retail** (bagged garden grade)
  - **H7–H10 → non-saleable** (reported, not harvested)

```{r band}
intervals <- horizons |>
  mutate(
    thickness_cm = depth_bot_cm - depth_top_cm,
    h_num = suppressWarnings(as.integer(sub("^H", "", von_post))),
    band  = dplyr::case_when(
      h_num %in% 1:4  ~ "professional",
      h_num %in% 5:6  ~ "retail",
      h_num %in% 7:10 ~ "non_saleable",
      TRUE            ~ NA_character_
    )
  )
```

## Apply the reclamation buffer (deepest 50 cm, by depth)

The reclamation allowance is the **deepest `reclaim_depth_cm` of the physical
column, whatever H values occupy it** — a contiguous basal layer left on the
mineral substrate. We do *not* assume the base is non-saleable: a non-monotonic
flip can put commercial peat at the bottom, and this geometry handles that.

For each interval we split its thickness into the part **above** the reclamation
line (harvestable) and the part **within** the buffer.

```{r reclamation}
recl <- params$reclaim_depth_cm

intervals <- intervals |>
  group_by(plot_id) |>
  mutate(
    mineral_depth_cm = max(depth_bot_cm, na.rm = TRUE),
    recl_line_cm     = mineral_depth_cm - recl,                 # boundary depth
    # portion of this interval shallower than the reclamation line:
    harvest_cm = pmax(0, pmin(depth_bot_cm, recl_line_cm) - depth_top_cm),
    buffer_cm  = thickness_cm - harvest_cm
  ) |>
  ungroup()
```

## Per-hole harvestable totals (total-resource interpretation)

Commercial peat is summed across the whole column regardless of vertical
position (a non-saleable layer above commercial peat is not treated as
sterilising it). The four quantities below partition the full peat column with
no double-counting:
`prof_harv + retail_harv + nonsaleable + commercial_reclaim = peat depth`.

```{r hole-summary}
hole_summary <- intervals |>
  group_by(plot_id) |>
  summarise(
    mineral_depth_cm = first(mineral_depth_cm),
    # harvestable = commercial-grade AND above the reclamation line
    prof_harv_cm   = sum(harvest_cm[band == "professional"], na.rm = TRUE),
    retail_harv_cm = sum(harvest_cm[band == "retail"],       na.rm = TRUE),
    # non-saleable = all H7–H10 anywhere in the column
    nonsaleable_cm = sum(thickness_cm[band == "non_saleable"], na.rm = TRUE),
    # commercial peat sacrificed to reclamation = H1–H6 within the basal buffer
    commercial_reclaim_cm = sum(buffer_cm[band %in% c("professional", "retail")],
                                na.rm = TRUE),
    .groups = "drop"
  ) |>
  # thicknesses in metres for volume integration
  mutate(
    prof_harv_m            = prof_harv_cm / 100,
    retail_harv_m          = retail_harv_cm / 100,
    nonsaleable_m          = nonsaleable_cm / 100,
    commercial_reclaim_m   = commercial_reclaim_cm / 100
  ) |>
  left_join(plots, by = "plot_id")

hole_summary |>
  select(plot_id, mineral_depth_cm, prof_harv_cm, retail_harv_cm,
         nonsaleable_cm, commercial_reclaim_cm)
```

## Add upland holes as 0-thickness control points

Test holes classed **Upland** have no peat profile, so they never appear in the
per-hole summary above. We add them back as explicit **0-thickness anchors** so
the interpolation drops toward zero at these known non-peat locations instead of
extrapolating peat into them.

```{r upland-anchors}
upland_anchors <- plots |>
  filter(awcs_wetland_class %in% "Upland", !plot_id %in% hole_summary$plot_id) |>
  mutate(mineral_depth_cm = 0,
         prof_harv_cm = 0, retail_harv_cm = 0,
         nonsaleable_cm = 0, commercial_reclaim_cm = 0,
         prof_harv_m = 0, retail_harv_m = 0,
         nonsaleable_m = 0, commercial_reclaim_m = 0)

hole_summary <- bind_rows(hole_summary, upland_anchors)

cat(nrow(upland_anchors), "upland 0-thickness anchor(s) added:",
    if (nrow(upland_anchors)) paste(upland_anchors$plot_id, collapse = ", ") else "none", "\n")
```

## Build spatial points (handles both UTM zones)

```{r points}
# Split by UTM zone so each gets the correct CRS, then reproject to a common one.
make_points <- function(df, zone, epsg) {
  sub <- df |> filter(utm_zone == zone)
  if (nrow(sub) == 0) return(NULL)
  st_as_sf(sub, coords = c("gps_easting", "gps_northing"), crs = epsg, remove = FALSE)
}
pts_11 <- make_points(hole_summary, "11N", 26911)   # NAD83 / UTM zone 11N
pts_12 <- make_points(hole_summary, "12N", 26912)   # NAD83 / UTM zone 12N

# Reproject everything to the CRS of the ownership boundary (the analysis CRS).
ownership <- st_read(here(params$ownership_boundary), quiet = TRUE)
approved  <- st_read(here(params$approved_boundary),  quiet = TRUE)
target_crs <- st_crs(ownership)

pts <- bind_rows(
  if (!is.null(pts_11)) st_transform(pts_11, target_crs),
  if (!is.null(pts_12)) st_transform(pts_12, target_crs)
)
approved <- st_transform(approved, target_crs)

pts_v       <- terra::vect(pts)
approved_v  <- terra::vect(approved)
ownership_v <- terra::vect(ownership)
```

## Resolve grid resolution from core spacing

The interpolation grid is sized from the data: mean nearest-neighbour spacing
between cores ÷ 10, clamped to 5–25 m. Resolution controls boundary-clipping
accuracy for the volume integral, not the interpolated values (those are set by
core density). Override by setting `raster_res_m` to a positive value.

```{r resolution}
nn_dist <- sf::st_distance(pts)            # pairwise distances (metres; CRS is UTM)
diag(nn_dist) <- NA
mean_nn <- mean(apply(nn_dist, 1, min, na.rm = TRUE))

res_m <- params$raster_res_m
if (is.null(res_m) || res_m <= 0) {
  res_m <- max(5, min(25, round(mean_nn / 10)))
}
cat(sprintf("Mean nearest-neighbour core spacing: %.0f m  ->  grid resolution: %d m\n",
            mean_nn, as.integer(res_m)))
```

## IDW interpolation over the area

```{r interpolate}
# Raster template spanning the ownership boundary (the larger extent).
template <- terra::rast(ownership_v, resolution = res_m)

# Auto radius = extent diagonal → every point contributes (global IDW).
radius <- params$idw_radius_m
if (is.null(radius) || radius <= 0) {
  e <- terra::ext(template)
  radius <- sqrt((e$xmax - e$xmin)^2 + (e$ymax - e$ymin)^2)
}

interp_field <- function(field) {
  terra::interpIDW(template, pts_v, field = field,
                   radius = radius, power = params$idw_power)
}

surf <- list(
  professional        = interp_field("prof_harv_m"),
  retail              = interp_field("retail_harv_m"),
  non_saleable        = interp_field("nonsaleable_m"),
  reclamation_loss    = interp_field("commercial_reclaim_m")
)

# Persist the surfaces for the visualization step.
for (nm in names(surf)) {
  terra::writeRaster(
    surf[[nm]],
    here(params$output_dir, paste0(params$project_id, "_thickness_", nm, ".tif")),
    overwrite = TRUE)
}
```

## Integrate volumes within each boundary

`volume (m³) = Σ (cell thickness in m × cell area in m²)` over the clipped surface.
CRS is metric UTM, so cell area is the product of the raster resolution.

```{r volumes}
volume_in <- function(surface, boundary_v) {
  m <- terra::mask(terra::crop(surface, boundary_v), boundary_v)
  cell_area <- prod(terra::res(m))                       # m² per cell
  terra::global(m, "sum", na.rm = TRUE)[1, 1] * cell_area
}

make_row <- function(label, surface) {
  data.frame(
    product      = label,
    approved_m3  = volume_in(surface, approved_v),
    ownership_m3 = volume_in(surface, ownership_v)
  )
}

volumes <- bind_rows(
  make_row("Professional peat (H1-H4)",        surf$professional),
  make_row("Retail peat (H5-H6)",              surf$retail),
  make_row("Non-saleable peat (H7-H10)",       surf$non_saleable),
  make_row("Commercial peat left to reclamation", surf$reclamation_loss)
) |>
  mutate(across(c(approved_m3, ownership_m3), ~ round(.x, 0)))

volumes
```

## Save

```{r save}
saveRDS(intervals,    here("data-raw", paste0(params$project_id, "_peat_transformed.rds")))
saveRDS(hole_summary, here("data-raw", paste0(params$project_id, "_peat_hole_summary.rds")))
saveRDS(volumes,      here("data-raw", paste0(params$project_id, "_peat_volumes.rds")))
write.csv(volumes,
          here("data-raw", paste0(params$project_id, "_peat_volumes.csv")),
          row.names = FALSE)
```
