---
title: "Peat Transformation, Volume & Spatial Interpolation"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  approved_boundary: "data-raw/projects/approved_harvest_area.shp"
  ownership_boundary: "data-raw/projects/ownership_boundary.shp"
  boundary_setback_m: 30    # new-approval harvest setback from the property boundary (expansion remainder only; the approved area already incorporates its setback)
  waterbody_setback_m: 45   # setback from waterbodies, applied everywhere incl. the approved lands (public-land default condition)
  waterbodies_boundary: ""  # optional; comma-separated paths. Polygons (lakes/ponds) AND polylines (creeks/streams) both get the setback. "" = not applied
  roads_boundary: ""        # optional roads (comma-separated ok); centerlines OR footprint polygons; "" = no infrastructure exclusion
  road_half_width_m: 6      # half the running width — buffers road CENTERLINES into the surface (12 m road -> 6 m); ignored for polygon footprints
  road_buffer_m: 5          # operational setback (m) beyond the road surface (ditch/shoulder/stability)
  utilities_boundary: ""    # optional; comma-separated paths to pipelines / utilities / well heads & pads (points, lines or polygons). "" = not applied
  utility_buffer_m: 30      # standard buffer (m) around all utilities
  reclaim_depth_cm: 50      # material left on the mineral substrate for reclamation
  raster_res_m: 0           # 0 = auto (mean nearest-neighbour core spacing / 10, clamped 5-25 m)
  idw_power: 2              # IDW distance-decay exponent
  idw_radius_m: 0           # 0 = auto (full extent diagonal → global IDW)
  output_dir: "reports/figures"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(sf)
library(terra)
library(here)
dir.create(here(params$output_dir), showWarnings = FALSE, recursive = TRUE)
```

## Load clean data

```{r load}
peat     <- readRDS(here("data-raw", paste0(params$project_id, "_peat_clean.rds")))
plots    <- peat$plots
horizons <- peat$horizons
```

## Band each interval by Von Post humification

Banding is **independent of vertical order**, so non-monotonic cores need no
special handling — each interval is assigned by its own H value.

  - **H1–H4 → professional** (premium horticultural grade)
  - **H5–H6 → retail** (bagged garden grade)
  - **H7–H10 → non-saleable** (reported, not harvested)

```{r band}
intervals <- horizons |>
  mutate(
    thickness_cm = depth_bot_cm - depth_top_cm,
    h_num = suppressWarnings(as.integer(sub("^H", "", von_post))),
    band  = dplyr::case_when(
      h_num %in% 1:4  ~ "professional",
      h_num %in% 5:6  ~ "retail",
      h_num %in% 7:10 ~ "non_saleable",
      TRUE            ~ NA_character_
    )
  )
```

## Apply the reclamation buffer (deepest 50 cm, by depth)

The reclamation allowance is the **deepest `reclaim_depth_cm` of the physical
column, whatever H values occupy it** — a contiguous basal layer left on the
mineral substrate. We do *not* assume the base is non-saleable: a non-monotonic
flip can put commercial peat at the bottom, and this geometry handles that.

For each interval we split its thickness into the part **above** the reclamation
line (harvestable) and the part **within** the buffer.

```{r reclamation}
recl <- params$reclaim_depth_cm

intervals <- intervals |>
  group_by(plot_id) |>
  mutate(
    mineral_depth_cm = max(depth_bot_cm, na.rm = TRUE),
    recl_line_cm     = mineral_depth_cm - recl,                 # boundary depth
    # portion of this interval shallower than the reclamation line:
    harvest_cm = pmax(0, pmin(depth_bot_cm, recl_line_cm) - depth_top_cm),
    buffer_cm  = thickness_cm - harvest_cm
  ) |>
  ungroup()
```

## Per-hole harvestable totals (total-resource interpretation)

Commercial peat is summed across the whole column regardless of vertical
position (a non-saleable layer above commercial peat is not treated as
sterilising it). The four quantities below partition the full peat column with
no double-counting:
`prof_harv + retail_harv + nonsaleable + commercial_reclaim = peat depth`.

```{r hole-summary}
hole_summary <- intervals |>
  group_by(plot_id) |>
  summarise(
    mineral_depth_cm = first(mineral_depth_cm),
    # harvestable = commercial-grade AND above the reclamation line
    prof_harv_cm   = sum(harvest_cm[band == "professional"], na.rm = TRUE),
    retail_harv_cm = sum(harvest_cm[band == "retail"],       na.rm = TRUE),
    # non-saleable = all H7–H10 anywhere in the column
    nonsaleable_cm = sum(thickness_cm[band == "non_saleable"], na.rm = TRUE),
    # commercial peat sacrificed to reclamation = H1–H6 within the basal buffer
    commercial_reclaim_cm = sum(buffer_cm[band %in% c("professional", "retail")],
                                na.rm = TRUE),
    .groups = "drop"
  ) |>
  # thicknesses in metres for volume integration
  mutate(
    prof_harv_m            = prof_harv_cm / 100,
    retail_harv_m          = retail_harv_cm / 100,
    nonsaleable_m          = nonsaleable_cm / 100,
    commercial_reclaim_m   = commercial_reclaim_cm / 100
  ) |>
  left_join(plots, by = "plot_id")

hole_summary |>
  select(plot_id, mineral_depth_cm, prof_harv_cm, retail_harv_cm,
         nonsaleable_cm, commercial_reclaim_cm)
```

## Add upland holes as 0-thickness control points

Test holes classed **Upland** have no peat profile, so they never appear in the
per-hole summary above. We add them back as explicit **0-thickness anchors** so
the interpolation drops toward zero at these known non-peat locations instead of
extrapolating peat into them.

```{r upland-anchors}
upland_anchors <- plots |>
  filter(awcs_wetland_class %in% "Upland", !plot_id %in% hole_summary$plot_id) |>
  mutate(mineral_depth_cm = 0,
         prof_harv_cm = 0, retail_harv_cm = 0,
         nonsaleable_cm = 0, commercial_reclaim_cm = 0,
         prof_harv_m = 0, retail_harv_m = 0,
         nonsaleable_m = 0, commercial_reclaim_m = 0)

hole_summary <- bind_rows(hole_summary, upland_anchors)

cat(nrow(upland_anchors), "upland 0-thickness anchor(s) added:",
    if (nrow(upland_anchors)) paste(upland_anchors$plot_id, collapse = ", ") else "none", "\n")
```

## Build spatial points (handles both UTM zones)

```{r points}
# Split by UTM zone so each gets the correct CRS, then reproject to a common one.
make_points <- function(df, zone, epsg) {
  sub <- df |> filter(utm_zone == zone)
  if (nrow(sub) == 0) return(NULL)
  st_as_sf(sub, coords = c("gps_easting", "gps_northing"), crs = epsg, remove = FALSE)
}
pts_11 <- make_points(hole_summary, "11N", 26911)   # NAD83 / UTM zone 11N
pts_12 <- make_points(hole_summary, "12N", 26912)   # NAD83 / UTM zone 12N

# Reproject everything to the CRS of the ownership boundary (the analysis CRS).
ownership <- st_read(here(params$ownership_boundary), quiet = TRUE)
approved  <- st_read(here(params$approved_boundary),  quiet = TRUE)
target_crs <- st_crs(ownership)

pts <- bind_rows(
  if (!is.null(pts_11)) st_transform(pts_11, target_crs),
  if (!is.null(pts_12)) st_transform(pts_12, target_crs)
)
approved <- st_transform(approved, target_crs)

pts_v       <- terra::vect(pts)
approved_v  <- terra::vect(approved)
ownership_v <- terra::vect(ownership)
```

## Resolve grid resolution from core spacing

The interpolation grid is sized from the data: mean nearest-neighbour spacing
between cores ÷ 10, clamped to 5–25 m. Resolution controls boundary-clipping
accuracy for the volume integral, not the interpolated values (those are set by
core density). Override by setting `raster_res_m` to a positive value.

```{r resolution}
nn_dist <- sf::st_distance(pts)            # pairwise distances (metres; CRS is UTM)
diag(nn_dist) <- NA
mean_nn <- mean(apply(nn_dist, 1, min, na.rm = TRUE))

res_m <- params$raster_res_m
if (is.null(res_m) || res_m <= 0) {
  res_m <- max(5, min(25, round(mean_nn / 10)))
}
cat(sprintf("Mean nearest-neighbour core spacing: %.0f m  ->  grid resolution: %d m\n",
            mean_nn, as.integer(res_m)))
```

## IDW interpolation over the area

```{r interpolate}
# Raster template spanning the ownership boundary (the larger extent).
template <- terra::rast(ownership_v, resolution = res_m)

# Auto radius = extent diagonal → every point contributes (global IDW).
radius <- params$idw_radius_m
if (is.null(radius) || radius <= 0) {
  e <- terra::ext(template)
  radius <- sqrt((e$xmax - e$xmin)^2 + (e$ymax - e$ymin)^2)
}

interp_field <- function(field) {
  terra::interpIDW(template, pts_v, field = field,
                   radius = radius, power = params$idw_power)
}

surf <- list(
  professional        = interp_field("prof_harv_m"),
  retail              = interp_field("retail_harv_m"),
  non_saleable        = interp_field("nonsaleable_m"),
  reclamation_loss    = interp_field("commercial_reclaim_m")
)

# Persist the surfaces for the visualization step.
for (nm in names(surf)) {
  terra::writeRaster(
    surf[[nm]],
    here(params$output_dir, paste0(params$project_id, "_thickness_", nm, ".tif")),
    overwrite = TRUE)
}
```

## Harvestable area: setbacks & infrastructure exclusions

Setbacks and existing infrastructure are applied **only here, at the integration
mask** — the IDW surfaces above are never altered. Peat is physically continuous
beneath roads and within setbacks; these are *harvestability* constraints, not
peat-absence. Each constraint is optional and a **no-op when its layer isn't
supplied**, so the pipeline still runs on projects without these data.

  - **Boundary setback** (`boundary_setback_m`, default 30 m): a new approval over
    the property remainder requires a setback from the property boundary. Applied
    to the **expansion remainder only** — the previously approved harvest area is
    assumed to already incorporate its setback, so it keeps its full extent.
  - **Waterbody setback** (`waterbody_setback_m`, default 45 m): applied
    **everywhere, including the approved lands** (conservative — any Water Act
    approval over the approved area may have expired). `st_buffer` handles both
    **polygon** waterbodies (lakes/ponds → buffered outward) and **polyline**
    waterbodies (creeks/streams → a corridor each side).
  - **Roads** (`roads_boundary` + `road_buffer_m`): the digitised footprint plus
    an operational buffer, removed as inaccessible/sterilised.

```{r harvestable-region}
# Buffer one or more (comma-separated) vector layers by `dist` and dissolve the
# result. Works for polygons AND polylines, so a single helper covers waterbodies
# (poly lakes + line creeks) and road footprints alike.
buffer_union <- function(paths_csv, dist) {
  if (is.null(paths_csv) || !nzchar(paths_csv)) return(NULL)
  parts <- trimws(strsplit(paths_csv, ",")[[1]]); parts <- parts[parts != ""]
  bufs <- lapply(parts, function(p) {
    if (!file.exists(here(p))) { warning("layer not found, skipped: ", p); return(NULL) }
    g <- sf::st_make_valid(sf::st_transform(sf::st_read(here(p), quiet = TRUE), target_crs))
    sf::st_union(sf::st_buffer(sf::st_geometry(g), dist))
  })
  bufs <- bufs[!vapply(bufs, is.null, logical(1))]
  if (!length(bufs)) return(NULL)
  out <- bufs[[1]]; for (b in bufs[-1]) out <- sf::st_union(out, b)
  sf::st_make_valid(out)
}

water_excl <- buffer_union(params$waterbodies_boundary, params$waterbody_setback_m)

# Roads are geometry-aware: LINESTRING centerlines are buffered by `half_width`
# to reconstruct the running surface (a 12 m road = 6 m half-width); POLYGON
# footprints already represent the surface and are left as-is. The operational
# `setback` (ditch/shoulder/stability) is then applied to whichever result.
road_footprint <- function(paths_csv, half_width, setback) {
  if (is.null(paths_csv) || !nzchar(paths_csv)) return(NULL)
  parts <- trimws(strsplit(paths_csv, ",")[[1]]); parts <- parts[parts != ""]
  geoms <- lapply(parts, function(p) {
    if (!file.exists(here(p))) { warning("layer not found, skipped: ", p); return(NULL) }
    g    <- sf::st_make_valid(sf::st_transform(sf::st_read(here(p), quiet = TRUE), target_crs))
    geom <- sf::st_geometry(g)
    is_line <- grepl("LINE", as.character(sf::st_geometry_type(g)))
    if (any(is_line)) geom[is_line] <- sf::st_buffer(geom[is_line], half_width)  # centerline -> surface
    sf::st_union(geom)
  })
  geoms <- geoms[!vapply(geoms, is.null, logical(1))]
  if (!length(geoms)) return(NULL)
  fp <- geoms[[1]]; for (b in geoms[-1]) fp <- sf::st_union(fp, b)
  fp <- sf::st_make_valid(fp)
  if (setback > 0) fp <- sf::st_make_valid(sf::st_buffer(fp, setback))           # operational setback
  fp
}
roads_excl <- road_footprint(params$roads_boundary, params$road_half_width_m, params$road_buffer_m)

# Utilities (pipelines, utility corridors, well heads & pads): a standard buffer
# around each, regardless of geometry — pads buffer outward, pipelines into a
# corridor each side, well-head points into a circle.
utils_excl <- buffer_union(params$utilities_boundary, params$utility_buffer_m)

# Harvestable regions (sf geometry):
#   approved  -> kept as-is (already net of its own boundary setback)
#   ownership -> approved (full extent) UNION the property shrunk by the boundary
#                setback, so the 30 m setback applies only to the expansion remainder.
expansion_sf   <- sf::st_make_valid(sf::st_buffer(ownership, -params$boundary_setback_m))
approved_base  <- sf::st_geometry(sf::st_make_valid(approved))
ownership_base <- sf::st_union(approved_base, sf::st_geometry(expansion_sf))

# Remove exclusions from a region. `include_water` lets us also build a variant
# WITHOUT the waterbody setback, so the waterbody's marginal contribution can be
# reported separately — it's the setback most likely to change or shrink, and the
# difference answers "how much harvestable peat would return if it did".
apply_excl <- function(region, include_water = TRUE) {
  region <- sf::st_make_valid(region)
  if (include_water && !is.null(water_excl)) region <- sf::st_difference(region, water_excl)
  if (!is.null(roads_excl)) region <- sf::st_difference(region, roads_excl)
  if (!is.null(utils_excl)) region <- sf::st_difference(region, utils_excl)
  sf::st_make_valid(region)
}
approved_harv          <- apply_excl(approved_base)
ownership_harv         <- apply_excl(ownership_base)
approved_harv_nowater  <- apply_excl(approved_base,  include_water = FALSE)
ownership_harv_nowater <- apply_excl(ownership_base, include_water = FALSE)

vv <- function(g) terra::vect(sf::st_sf(geometry = g))
approved_harv_v          <- vv(approved_harv)
ownership_harv_v         <- vv(ownership_harv)
approved_harv_nowater_v  <- vv(approved_harv_nowater)
ownership_harv_nowater_v <- vv(ownership_harv_nowater)

# Buffered exclusion ZONES, labelled by layer, for the comparison maps in 04.
# The boundary-setback zone is the perimeter strip removed from the expansion
# remainder; the others are the buffers themselves. Clipped to the study area
# for display only — the volume math above differences the unclipped buffers
# against each boundary.
own_geom  <- sf::st_union(sf::st_geometry(sf::st_make_valid(ownership)))
appr_geom <- sf::st_union(sf::st_geometry(sf::st_make_valid(approved)))
setback_zone <- sf::st_difference(sf::st_difference(own_geom, appr_geom),
                                  sf::st_geometry(expansion_sf))

zone <- function(geom, key, label) {
  if (is.null(geom) || !length(geom) || all(sf::st_is_empty(geom))) return(NULL)
  g <- suppressWarnings(sf::st_intersection(sf::st_make_valid(sf::st_geometry(geom)), own_geom))
  if (!length(g) || all(sf::st_is_empty(g))) return(NULL)
  sf::st_sf(key = key, type = label,
            geometry = sf::st_cast(sf::st_union(g), "MULTIPOLYGON"))
}
exclusion_zones <- do.call(rbind, Filter(Negate(is.null), list(
  zone(setback_zone, "boundary",  paste0("Boundary setback (", params$boundary_setback_m, " m)")),
  zone(water_excl,   "waterbody", paste0("Waterbody setback (", params$waterbody_setback_m, " m)")),
  zone(roads_excl,   "road",      "Road exclusion (footprint)"),
  zone(utils_excl,   "utility",   paste0("Utilities buffer (", params$utility_buffer_m, " m)"))
)))

# Honest record of what was actually applied in this run (drives the assumptions
# section in 04_visualize — never claim a deduction the numbers don't reflect).
exclusions_applied <- data.frame(
  exclusion = c("Boundary setback (expansion remainder)",
                "Waterbody setback (everywhere)",
                "Road exclusion",
                "Utilities exclusion (pipelines, utilities, well heads/pads)"),
  setting   = c(paste0(params$boundary_setback_m, " m"),
                paste0(params$waterbody_setback_m, " m"),
                paste0(2 * params$road_half_width_m, " m surface (centerlines) + ",
                       params$road_buffer_m, " m setback"),
                paste0(params$utility_buffer_m, " m")),
  applied   = c(params$boundary_setback_m > 0, !is.null(water_excl),
                !is.null(roads_excl), !is.null(utils_excl))
)
exclusions_applied
```

## Integrate volumes within each boundary

`volume (m³) = Σ (cell thickness in m × cell area in m²)` over the clipped surface.
CRS is metric UTM, so cell area is the product of the raster resolution. The
**in-place** table integrates over the full boundaries (the geological resource);
the **harvestable** table integrates the commercial surfaces over the harvestable
region (boundary minus setbacks minus infrastructure).

```{r volumes}
# Area-weighted integration: rasterize the region to FRACTIONAL cell coverage
# (cover = TRUE) rather than a cell-centre mask. This is essential for narrow
# exclusions — a 12 m road or a thin setback strip is sub-pixel at the ~25 m
# grid, so a centre-based mask would drop no cells and silently ignore the
# deduction. Coverage weighting captures partial cells proportionally, so the
# result is resolution-robust and the harvestable + excluded split reconciles
# exactly to the in-place total.
volume_in <- function(surface, region_v) {
  cov <- terra::rasterize(region_v, surface, cover = TRUE)
  cell_area <- prod(terra::res(surface))                 # m² per cell
  terra::global(surface * cov, "sum", na.rm = TRUE)[1, 1] * cell_area
}

make_row <- function(label, surface) {
  data.frame(
    product      = label,
    approved_m3  = volume_in(surface, approved_v),
    ownership_m3 = volume_in(surface, ownership_v)
  )
}

# In-place resource over the full boundaries (the geological partition).
volumes <- bind_rows(
  make_row("Professional peat (H1-H4)",        surf$professional),
  make_row("Retail peat (H5-H6)",              surf$retail),
  make_row("Non-saleable peat (H7-H10)",       surf$non_saleable),
  make_row("Commercial peat left to reclamation", surf$reclamation_loss)
) |>
  mutate(across(c(approved_m3, ownership_m3), ~ round(.x, 0)))

volumes

# Net harvestable commercial peat = professional/retail surfaces integrated over
# the harvestable region. The excluded line = commercial in-place minus net, so
# the harvestable + excluded partition reconciles against the in-place table.
harv_row <- function(label, surface) {
  data.frame(
    product      = label,
    approved_m3  = volume_in(surface, approved_harv_v),
    ownership_m3 = volume_in(surface, ownership_harv_v)
  )
}
commercial_gross <- function(b_v) volume_in(surf$professional, b_v) + volume_in(surf$retail, b_v)
commercial_net   <- function(h_v) volume_in(surf$professional, h_v) + volume_in(surf$retail, h_v)

# Excluded volume is split so the waterbody setback (most likely to change) is
# isolated from the rest. Waterbody = its MARGINAL contribution (extra peat
# removed once everything else is already excluded); other = everything but water.
# Net + waterbody-excluded + other-excluded reconciles to commercial in-place.
harvestable <- bind_rows(
  harv_row("Professional peat — harvestable", surf$professional),
  harv_row("Retail peat — harvestable",       surf$retail),
  data.frame(
    product      = "Commercial peat excluded — waterbody setback",
    approved_m3  = commercial_net(approved_harv_nowater_v)  - commercial_net(approved_harv_v),
    ownership_m3 = commercial_net(ownership_harv_nowater_v) - commercial_net(ownership_harv_v)
  ),
  data.frame(
    product      = "Commercial peat excluded — other setbacks & infrastructure",
    approved_m3  = commercial_gross(approved_v)  - commercial_net(approved_harv_nowater_v),
    ownership_m3 = commercial_gross(ownership_v) - commercial_net(ownership_harv_nowater_v)
  )
) |>
  mutate(across(c(approved_m3, ownership_m3), ~ round(.x, 0)))

harvestable
```

## Save

```{r save}
saveRDS(intervals,    here("data-raw", paste0(params$project_id, "_peat_transformed.rds")))
saveRDS(hole_summary, here("data-raw", paste0(params$project_id, "_peat_hole_summary.rds")))
saveRDS(volumes,      here("data-raw", paste0(params$project_id, "_peat_volumes.rds")))
write.csv(volumes,
          here("data-raw", paste0(params$project_id, "_peat_volumes.csv")),
          row.names = FALSE)

saveRDS(harvestable,        here("data-raw", paste0(params$project_id, "_peat_harvestable.rds")))
saveRDS(exclusions_applied, here("data-raw", paste0(params$project_id, "_peat_exclusions.rds")))
if (!is.null(exclusion_zones))
  saveRDS(exclusion_zones,  here("data-raw", paste0(params$project_id, "_peat_exclusion_zones.rds")))

# Grid settings + harvestable regions, so 05_uncertainty reuses the exact same
# interpolation grid and clip geometry (single source of truth).
saveRDS(list(res_m = res_m, radius = radius, idw_power = params$idw_power,
             approved_harv = approved_harv, ownership_harv = ownership_harv),
        here("data-raw", paste0(params$project_id, "_peat_geom.rds")))
write.csv(harvestable,
          here("data-raw", paste0(params$project_id, "_peat_harvestable.csv")),
          row.names = FALSE)
```
