---
title: "Peat Visualization & Volume Tables"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  approved_boundary: "data-raw/projects/approved_harvest_area.shp"
  ownership_boundary: "data-raw/projects/ownership_boundary.shp"
  roads_boundary: ""       # optional; drawn on the thickness maps (comma-separated paths ok)
  waterbodies_boundary: "" # optional; polygons and/or lines drawn on the thickness maps
  boundary_setback_m: 30   # for the assumptions text — must match the value used in 03_transform
  waterbody_setback_m: 45  # for the assumptions text — must match 03_transform
  road_half_width_m: 6     # for the assumptions text — must match 03_transform (12 m road = 6 m half-width)
  road_buffer_m: 5         # for the assumptions text — must match 03_transform
  utility_buffer_m: 30     # for the assumptions text — must match 03_transform
  basemap: false           # TRUE to fetch an Esri World Imagery backdrop (needs maptiles + internet)
  section_holes: ""        # ordered transect for the single cross-section; empty = skip (per-project, e.g. "PL05, PL04, PL01")
  output_dir: "reports/figures"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 6, dpi = 300)
library(dplyr)
library(ggplot2)
library(sf)
library(terra)
library(here)
# Shared grade palette / legend labels / classify_grade() — single source of
# truth, also used by analysis/peat/cross_sections.R (appendix fence diagrams).
source(here("analysis/peat/grade_palette.R"))
# zone_epsg() / project_holes(): zone-safe hole points in one common CRS.
source(here("analysis/peat/spatial_helpers.R"))
dir.create(here(params$output_dir), showWarnings = FALSE, recursive = TRUE)
```

## Load transformed outputs

```{r load}
intervals    <- readRDS(here("data-raw", paste0(params$project_id, "_peat_transformed.rds")))
hole_summary <- readRDS(here("data-raw", paste0(params$project_id, "_peat_hole_summary.rds")))
volumes      <- readRDS(here("data-raw", paste0(params$project_id, "_peat_volumes.rds")))

ownership <- st_read(here(params$ownership_boundary), quiet = TRUE)
approved  <- st_transform(st_read(here(params$approved_boundary), quiet = TRUE),
                          st_crs(ownership))

# Test-hole points in the analysis (ownership) CRS, for overlaying on the maps.
# Each hole is built in its recorded UTM zone, then reprojected; proj_x/proj_y
# give zone-safe coordinates. Zone read per hole from utm_zone.
pts <- project_holes(hole_summary, st_crs(ownership))
pts_v <- terra::vect(pts)

# Optional overlay layers for the maps (roads, waterbodies). Each param accepts
# one or more comma-separated shapefiles; we pull out geometries by type so they
# can be styled differently (waterbody polygons filled, lines drawn as lines).
# Returns a terra SpatVector, or NULL when nothing matches (so the maps still
# render on projects without these data).
gather_v <- function(paths_csv, want = NULL) {
  if (is.null(paths_csv) || !nzchar(paths_csv)) return(NULL)
  parts <- trimws(strsplit(paths_csv, ",")[[1]]); parts <- parts[parts != ""]
  geoms <- list()
  for (p in parts) {
    if (!file.exists(here(p))) next
    g  <- st_transform(st_read(here(p), quiet = TRUE), st_crs(ownership))
    gt <- as.character(st_geometry_type(g))
    keep <- if (is.null(want)) rep(TRUE, length(gt)) else grepl(want, gt)
    if (any(keep)) geoms[[length(geoms) + 1]] <- st_geometry(g)[keep]
  }
  if (!length(geoms)) return(NULL)
  terra::vect(st_sf(geometry = do.call(c, geoms)))
}
roads_v      <- gather_v(params$roads_boundary,      "LINE")
water_poly_v <- gather_v(params$waterbodies_boundary, "POLY")
water_line_v <- gather_v(params$waterbodies_boundary, "LINE")

# Buffered exclusion zones (saved by 03_transform), labelled by layer, for the
# setback-comparison map set. Colours keyed to the associated layer.
zones_path <- here("data-raw", paste0(params$project_id, "_peat_exclusion_zones.rds"))
exclusion_zones <- if (file.exists(zones_path)) readRDS(zones_path) else NULL
zone_cols <- c(boundary = "#7570b3", waterbody = "#08519c",
               road = "#543005", utility = "#e6550d")
```

## Per-hole field summary

```{r tbl-field-summary}
# One row per test hole, surfacing the per-hole field measurements the client
# (Sun Gro) asked for: AWCS wetland class, dominant surface bryophyte, depth to
# water table, depth to mineral contact (peat depth), pH and EC — plus the
# harvestable thickness split by product grade (professional H1-H4, retail
# H5-H6, each net of the 0.5 m basal reclamation allowance). All columns already
# live in hole_summary (intervals + the joined plots sheet); this just presents
# them. NA-safe: frost/disturbance can leave individual fields unrecorded, and
# Upland anchor holes carry 0 peat depth.
num <- function(x, d) ifelse(is.na(x), "—",
  formatC(round(as.numeric(x), d), format = "f", digits = d, big.mark = ","))
chr <- function(x) ifelse(is.na(x) | !nzchar(as.character(x)), "—", as.character(x))

field_tbl <- hole_summary |>
  arrange(plot_id) |>
  transmute(
    `Test Hole`              = plot_id,
    `Wetland Class (AWCS)`   = chr(awcs_wetland_class),
    `Dominant Bryophyte`     = chr(dominant_bryophyte),
    `Peat Depth (m)`         = num(mineral_depth_cm / 100, 2),
    `Water Table (m)`        = num(water_table_depth_m, 2),
    `pH`                     = num(ph, 1),
    `EC (µS/cm)`             = num(conductivity_us_cm, 0),
    `Professional H1-H4 (m)` = num(prof_harv_m, 2),
    `Retail H5-H6 (m)`       = num(retail_harv_m, 2)
  )

# Save a report-ready CSV alongside the other pipeline outputs.
write.csv(field_tbl,
          here("data-raw", paste0(params$project_id, "_peat_field_summary.csv")),
          row.names = FALSE, fileEncoding = "UTF-8")

knitr::kable(
  field_tbl,
  align   = c("l", "l", "l", "r", "r", "r", "r", "r", "r"),
  caption = paste0(
    "Per-hole field summary — ", params$project_id,
    ". One row per test hole: AWCS wetland classification, dominant surface ",
    "bryophyte, depth to mineral contact (peat depth), depth to water table, pH, ",
    "and electrical conductivity, with harvestable thickness by product grade ",
    "(professional H1–H4, retail H5–H6) net of the 0.5 m basal reclamation ",
    "allowance. “—” = not recorded in the field (e.g. frozen ground) or ",
    "not applicable (Upland anchor holes carry 0 m peat). Harvestable thicknesses ",
    "are the per-hole values interpolated for the volume estimates."))
```

## Volume summary table

```{r tbl-volumes}
knitr::kable(
  volumes,
  col.names = c("Product", "Approved area (m³)", "Ownership area (m³)"),
  format.args = list(big.mark = ","),
  caption = paste0("In-place peat resource — ", params$project_id,
                   ". The geological partition over the full boundaries; ",
                   "commercial grades are net of the reclamation allowance only."))
```

_The net harvestable peat table — with bootstrap uncertainty intervals on the
harvestable rows, the waterbody/other excluded split, and accompanying notes — is
produced in `05_uncertainty.Rmd` so it carries its uncertainty range. See that
step's output for the report-ready table._

## Assumptions and limitations

```{r assumptions, results='asis'}
excl_path <- here("data-raw", paste0(params$project_id, "_peat_exclusions.rds"))

cat(sprintf("
**Harvest setbacks.** Reported harvestable volumes exclude regulatory and operational setbacks as follows:

- The **previously approved harvest area** is assumed to already incorporate its required boundary setback; no additional boundary setback is applied to it.
- The **expansion area** (remainder of the property, which would require a new approval) is assumed to require a **%d m harvest setback** from the property boundary.
- A **%d m setback from waterbodies** is applied throughout the study area, **including the previously approved lands** — applied conservatively because any *Water Act* approval over the approved lands may have expired and require updating under current conditions. Both polygon waterbodies (lakes/ponds) and polyline waterbodies (creeks/permanent streams) receive this setback.

These setback distances reflect the conditions applicable to **public land**. If the property is **private**, the applicable setbacks may differ; this can be amended in a future report if required.

**Existing infrastructure.** Peat beneath existing roads (constructed as fill/corduroy over intact peat) is present but not harvestable. Road centrelines are buffered to the **%d m** running width, plus a **%d m** operational setback (ditch/shoulder/stability), and that footprint is excluded from harvestable volume.

**Utilities.** A standard **%d m** buffer is applied around all utilities — pipelines, utility corridors, and well heads/pads — and the buffered area is excluded from harvestable volume.

**Method note.** Setbacks and infrastructure are applied only when integrating volume; the interpolated peat-thickness surfaces are not altered (peat is physically continuous beneath roads and within setbacks). Volumes are integrated by fractional cell coverage, so exclusions narrower than the grid cell (e.g. a road corridor) are captured proportionally rather than missed.
",
  as.integer(params$boundary_setback_m),
  as.integer(params$waterbody_setback_m),
  as.integer(2 * params$road_half_width_m),
  as.integer(params$road_buffer_m),
  as.integer(params$utility_buffer_m)))

if (file.exists(excl_path)) {
  ex <- readRDS(excl_path)
  ex$applied <- ifelse(ex$applied, "applied", "**not applied — no layer supplied**")
  cat("\n\n**Exclusions applied in this run** (the figures above reflect only the rows marked *applied*):\n\n")
  print(knitr::kable(ex, col.names = c("Exclusion", "Setting", "Status")))
}
```

## Per-hole peat profiles by Von Post humification

```{r fig-profiles, fig.height=6}
# Grade palette, levels, labels and classify_grade() come from grade_palette.R
# (sourced in setup) so this figure shares one legend key with the cross-sections.
p_prof <- intervals |>
  classify_grade() |>
  ggplot(aes(x = plot_id,
             ymin = -depth_bot_cm, ymax = -depth_top_cm,
             colour = grade)) +
  geom_linerange(linewidth = 8) +
  scale_colour_manual(values = band_cols, name = "Layer", drop = FALSE,
                      labels = grade_labels) +
  scale_y_continuous(labels = ~ abs(.)) +
  labs(title = "Peat Profiles by Grade", x = "Test hole", y = "Depth (cm)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_prof
ggsave(here(params$output_dir, paste0(params$project_id, "_peat_profiles.png")),
       p_prof, width = 9, height = 5, dpi = 300)
```

## Per-hole peat profiles with the reclamation buffer

The same profiles with a **dashed red line** at the base of the harvestable
column (mineral depth − reclamation allowance). Everything below the line is the
deepest material left in place for reclamation. Provided as a separate figure so
it can be compared against the plain profiles above.

```{r fig-profiles-reclamation, fig.height=6}
reclaim_cm  <- round(stats::median(intervals$mineral_depth_cm - intervals$recl_line_cm))
plot_levels <- sort(unique(intervals$plot_id))
recl_label  <- paste0(reclaim_cm, " cm reclamation buffer")
recl_lines  <- intervals |>
  distinct(plot_id, recl_line_cm) |>
  mutate(xpos = match(plot_id, plot_levels),
         y    = -pmax(recl_line_cm, 0))   # clamp shallow holes to the surface

# Map the line to a linetype aesthetic so it gets its own legend entry (drawn
# red via override.aes), rather than explaining it in a subtitle.
p_prof_recl <- p_prof +
  geom_segment(data = recl_lines, inherit.aes = FALSE,
               aes(x = xpos - 0.45, xend = xpos + 0.45, y = y, yend = y,
                   linetype = recl_label),
               colour = "red", linewidth = 0.7) +
  scale_linetype_manual(name = NULL, values = setNames("dashed", recl_label),
                        guide = guide_legend(override.aes = list(colour = "red"))) +
  labs(title = "Peat Profiles by Grade with Reclamation Buffer")

p_prof_recl
ggsave(here(params$output_dir, paste0(params$project_id, "_peat_profiles_reclamation.png")),
       p_prof_recl, width = 9, height = 5, dpi = 300)
```

## Cross-section (stratigraphic fence of logged test holes)

A vertical section along a chosen transect of test holes (`params$section_holes`).
Each hole is drawn as its **real logged grade column** at its true along-transect
distance; the surface (0) and the mineral base are connected between holes. The
ground between holes is left blank — this is honest logged stratigraphy, not
depth-resolved interpolation (Von Post flips with depth, so grade-at-depth is not
interpolated; see DESIGN).

```{r fig-cross-section, fig.height=5}
sec_ids <- trimws(strsplit(params$section_holes, ",")[[1]])
sec_ids <- sec_ids[sec_ids != ""]

# The transect is project-specific. With no section_holes supplied, skip this
# single cross-section (the appendix A–I series from cross_sections.R is the
# primary cross-section output); a named-but-missing hole is still a hard error.
make_section <- length(sec_ids) > 0
```

```{r fig-cross-section-build, fig.height=5, eval=make_section}
# Chosen holes in transect order, with coordinates, depth and water table. Use
# the common-CRS coordinates (proj_x/proj_y) so distances are zone-safe.
sec_xy <- sf::st_drop_geometry(pts) |>
  distinct(plot_id, proj_x, proj_y, mineral_depth_cm, water_table_depth_m)
sec_xy <- sec_xy[match(sec_ids, sec_xy$plot_id), ]
if (anyNA(sec_xy$proj_x))
  stop("section_holes not found in data: ",
       paste(sec_ids[is.na(sec_xy$proj_x)], collapse = ", "))

# Cumulative straight-line distance along the polyline through the holes (m).
sec_xy$dist_m <- cumsum(c(0, sqrt(diff(sec_xy$proj_x)^2 +
                                  diff(sec_xy$proj_y)^2)))
sec_xy <- left_join(sec_xy, distinct(intervals, plot_id, recl_line_cm), by = "plot_id")
sec_xy$wt_cm <- sec_xy$water_table_depth_m * 100   # water-table depth (cm)

# Real logged intervals for those holes, classified to the shared grade palette.
sec_intervals <- intervals |>
  filter(plot_id %in% sec_ids) |>
  classify_grade() |>
  left_join(sec_xy[, c("plot_id", "dist_m")], by = "plot_id")

bar_w   <- 0.012 * max(sec_xy$dist_m)   # core column half-width (m)
ann_lvls <- c("Mineral base", "50 cm reclamation buffer", "Water table")

p_section <- ggplot() +
  geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.4) +          # surface
  geom_rect(data = sec_intervals,                                           # logged cores
            aes(xmin = dist_m - bar_w, xmax = dist_m + bar_w,
                ymin = -depth_bot_cm, ymax = -depth_top_cm, fill = grade)) +
  geom_line(data = sec_xy, aes(x = dist_m, y = -mineral_depth_cm,           # mineral base
            colour = "Mineral base"), linetype = "dashed") +
  geom_line(data = sec_xy, aes(x = dist_m, y = -recl_line_cm,               # reclamation buffer
            colour = "50 cm reclamation buffer"), linetype = "dashed") +
  geom_line(data = sec_xy, aes(x = dist_m, y = -wt_cm,                      # water table
            colour = "Water table"), linetype = "dashed") +
  geom_text(data = sec_xy, aes(x = dist_m, y = 20, label = plot_id),
            angle = 90, hjust = 0, size = 2.6) +
  scale_fill_manual(values = band_cols, name = "Layer", drop = FALSE,
                    labels = grade_labels) +
  scale_colour_manual(name = NULL, breaks = ann_lvls,
    values = c("Mineral base" = "grey40",
               "50 cm reclamation buffer" = "red",
               "Water table" = "dodgerblue3"),
    guide = guide_legend(order = 2, override.aes = list(linetype = "dashed"))) +
  scale_y_continuous(labels = ~ abs(.), expand = expansion(mult = c(0.04, 0.22))) +
  labs(title = "Peat Cross-Section (logged test holes along transect)",
       subtitle = paste(sec_ids, collapse = " → "),
       x = "Distance along transect (m)", y = "Depth (cm)") +
  theme_bw() +
  guides(fill = guide_legend(order = 1))

# Inset location map: study area, all holes, and the transect highlighted.
transect_line <- sf::st_sfc(
  sf::st_linestring(as.matrix(sec_xy[, c("proj_x", "proj_y")])),
  crs = sf::st_crs(ownership))
sec_pts <- pts |> filter(plot_id %in% sec_ids)

inset <- ggplot() +
  geom_sf(data = ownership, fill = "grey92", colour = "grey55", linewidth = 0.3) +
  geom_sf(data = pts, colour = "grey55", size = 0.5) +
  geom_sf(data = transect_line, colour = "black", linewidth = 0.7) +
  geom_sf(data = sec_pts, colour = "red", size = 1.1) +
  theme_void() +
  theme(panel.background = element_rect(fill = "white", colour = "grey40"),
        plot.margin = margin(1, 1, 1, 1))

# Draw the section, then overlay the inset in the empty lower-left via a grid
# viewport (print.ggplot uses newpage = is.null(vp), so it won't clear the page).
draw_section <- function() {
  print(p_section)
  print(inset, vp = grid::viewport(x = 0.255, y = 0.34, width = 0.26, height = 0.30))
}
png(here(params$output_dir, paste0(params$project_id, "_peat_cross_section.png")),
    width = 10, height = 5, units = "in", res = 300)
draw_section()
dev.off()
draw_section()
```

## Harvestable thickness per hole

```{r fig-thickness}
thick_long <- hole_summary |>
  select(plot_id, Professional = prof_harv_cm, Retail = retail_harv_cm) |>
  tidyr::pivot_longer(-plot_id, names_to = "grade", values_to = "thickness_cm")

p_thick <- thick_long |>
  ggplot(aes(x = reorder(plot_id, -thickness_cm), y = thickness_cm, fill = grade)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = harvest_cols) +
  labs(title = "Harvestable Peat Thickness by Hole (net of reclamation)",
       x = "Test hole", y = "Thickness (cm)", fill = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_thick
ggsave(here(params$output_dir, paste0(params$project_id, "_peat_thickness.png")),
       p_thick, width = 9, height = 5, dpi = 300)
```

## Interpolated thickness maps with boundaries

```{r fig-maps, fig.height=6}
# Read back the surfaces written in 03_transform.
surf_path <- function(nm)
  here(params$output_dir, paste0(params$project_id, "_thickness_", nm, ".tif"))

plot_surface <- function(nm, title, contour = FALSE, show_setbacks = FALSE) {
  r <- terra::rast(surf_path(nm))
  r <- terra::mask(r, terra::vect(ownership))           # clip to study area

  # Contour levels are derived from the raster's own pretty breaks, so the
  # interval always matches the colour-scale legend (~0.5 m for professional,
  # ~0.1 m for retail, etc.) without hard-coding — adapts to any future surface.
  # The actual interval is reflected in the title.
  if (contour) {
    lev  <- pretty(range(terra::values(r), na.rm = TRUE))
    lev  <- lev[lev > 0]
    step <- if (length(lev) > 1) signif(lev[2] - lev[1], 3) else NA
    title <- paste0(title, " - ", step, " m contours")
  }

  # Optional Esri World Imagery backdrop (licensed for reports with attribution).
  if (isTRUE(params$basemap) && requireNamespace("maptiles", quietly = TRUE)) {
    bm <- maptiles::get_tiles(terra::ext(r), provider = "Esri.WorldImagery",
                              crop = TRUE, project = TRUE)
    terra::plotRGB(bm); terra::plot(r, alpha = 0.6, add = TRUE)
  } else {
    terra::plot(r, main = title, xlab = "Easting (m)", ylab = "Northing (m)",
                colNA = "grey85",                       # area outside the study boundary
                plg = list(title = "Peat Depth (m)", title.cex = 0.8))
  }
  terra::plot(terra::vect(ownership), border = "grey20", lwd = 1, add = TRUE)
  terra::plot(terra::vect(approved),  border = "red",    lwd = 2, add = TRUE)

  # Setback/exclusion ZONES (setbacks map set only): the buffered areas actually
  # removed from harvestable volume, filled semi-transparent and coloured by the
  # layer they belong to. Drawn BEFORE the contours so the depth contours stay
  # crisp on top of the zone fills.
  if (show_setbacks && !is.null(exclusion_zones)) {
    for (i in seq_len(nrow(exclusion_zones))) {
      col_i <- zone_cols[[exclusion_zones$key[i]]]
      terra::plot(terra::vect(exclusion_zones[i, ]),
                  col = adjustcolor(col_i, 0.40), border = col_i, lwd = 1, add = TRUE)
    }
  }

  if (contour) {
    # White lines and labels read well across the viridis range (low dark to high
    # bright); only the brightest yellow caps are marginal, and the haloed
    # plot_id point labels remain the primary reference markers.
    terra::contour(r, add = TRUE, levels = lev, col = "white",
                   lwd = 0.8, labcex = 0.7, drawlabels = TRUE)
  }

  # Waterbody and road overlays, drawn beneath the test-hole markers. These are
  # the features themselves; the harvestable-volume exclusions buffer them
  # (45 m waterbodies, 12 m roads) in 03_transform.
  if (!is.null(water_poly_v))
    terra::plot(water_poly_v, col = adjustcolor("#4292c6", 0.55),
                border = "#08519c", lwd = 1, add = TRUE)
  if (!is.null(water_line_v)) terra::lines(water_line_v, col = "#08519c", lwd = 1.6)
  if (!is.null(roads_v))      terra::lines(roads_v,      col = "#543005", lwd = 2)

  # Test holes, labelled with plot_id so the map ties back to the profile chart.
  terra::plot(pts_v, add = TRUE, col = "black", pch = 21, bg = "white", cex = 1.1)
  terra::text(pts_v, labels = "plot_id", pos = 3, cex = 0.55,
              halo = TRUE, hc = "white", hw = 0.2)

  # Legend (the red / grey outlines plus any overlays). Use terra::add_legend,
  # which targets the map panel — base legend() does not survive terra's
  # colour-bar layout and renders into the wrong region. Built dynamically so it
  # only lists overlays that were actually supplied.
  leg <- data.frame(
    lab = c("Reno Exploration Lands", "Tactful Developments Ltd. Boundary"),
    col = c("red", "grey20"), lwd = c(2, 1), lty = c(1, 1))
  if (!is.null(roads_v))
    leg <- rbind(leg, data.frame(lab = "Roads", col = "#543005", lwd = 2, lty = 1))
  if (!is.null(water_poly_v) || !is.null(water_line_v))
    leg <- rbind(leg, data.frame(lab = "Waterbodies", col = "#08519c", lwd = 2, lty = 1))
  if (show_setbacks && !is.null(exclusion_zones))
    leg <- rbind(leg, data.frame(lab = exclusion_zones$type,
                                 col = zone_cols[exclusion_zones$key], lwd = 6, lty = 1))
  terra::add_legend("bottomleft", legend = leg$lab, col = leg$col,
    lwd = leg$lwd, lty = leg$lty, bg = "white", box.col = "grey70", cex = 0.7)
}

save_map <- function(nm, title, suffix = "", contour = FALSE, show_setbacks = FALSE) {
  png(here(params$output_dir,
           paste0(params$project_id, "_map_", nm, suffix, ".png")),
      width = 1800, height = 1500, res = 200)
  plot_surface(nm, title, contour = contour, show_setbacks = show_setbacks)
  dev.off()
}

# Plain thickness maps (now with Easting/Northing axes and a legend title)
save_map("professional", "Professional peat thickness (m)")
save_map("retail",       "Retail peat thickness (m)")

# Second set: the same surfaces with contour lines (interval auto-set per surface)
save_map("professional", "Professional peat thickness (m)",
         suffix = "_contour", contour = TRUE)
save_map("retail",       "Retail peat thickness (m)",
         suffix = "_contour", contour = TRUE)

# Third set: setback / exclusion zones overlaid, coloured by layer, WITH peat
# depth contours, for direct comparison against the plain maps above.
save_map("professional", "Professional peat thickness with setbacks (m)",
         suffix = "_setbacks", show_setbacks = TRUE, contour = TRUE)
save_map("retail",       "Retail peat thickness with setbacks (m)",
         suffix = "_setbacks", show_setbacks = TRUE, contour = TRUE)

# Show inline as well
plot_surface("professional", "Professional peat thickness (m)")
plot_surface("retail",       "Retail peat thickness (m)")
plot_surface("professional", "Professional peat thickness with setbacks (m)",
             show_setbacks = TRUE, contour = TRUE)
plot_surface("retail",       "Retail peat thickness with setbacks (m)",
             show_setbacks = TRUE, contour = TRUE)
plot_surface("professional", "Professional peat thickness (m)", contour = TRUE)
plot_surface("retail",       "Retail peat thickness (m)", contour = TRUE)
```
