---
title: "Peat Visualization & Volume Tables"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  approved_boundary: "data-raw/projects/approved_harvest_area.shp"
  ownership_boundary: "data-raw/projects/ownership_boundary.shp"
  basemap: false           # TRUE to fetch an Esri World Imagery backdrop (needs maptiles + internet)
  section_holes: "PL17, PL16, PL15, PL21, PL09, PL04, PL05, PL01"  # ordered transect for the cross-section
  output_dir: "reports/figures"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 6, dpi = 300)
library(dplyr)
library(ggplot2)
library(sf)
library(terra)
library(here)
dir.create(here(params$output_dir), showWarnings = FALSE, recursive = TRUE)
```

## Load transformed outputs

```{r load}
intervals    <- readRDS(here("data-raw", paste0(params$project_id, "_peat_transformed.rds")))
hole_summary <- readRDS(here("data-raw", paste0(params$project_id, "_peat_hole_summary.rds")))
volumes      <- readRDS(here("data-raw", paste0(params$project_id, "_peat_volumes.rds")))

ownership <- st_read(here(params$ownership_boundary), quiet = TRUE)
approved  <- st_transform(st_read(here(params$approved_boundary), quiet = TRUE),
                          st_crs(ownership))

# Test-hole points in the analysis (ownership) CRS, for overlaying on the maps.
# Split by UTM zone so each gets the right CRS, then reproject to the map CRS.
make_points <- function(df, zone, epsg) {
  sub <- dplyr::filter(df, utm_zone == zone)
  if (nrow(sub) == 0) return(NULL)
  st_as_sf(sub, coords = c("gps_easting", "gps_northing"), crs = epsg, remove = FALSE)
}
pts <- dplyr::bind_rows(
  { p <- make_points(hole_summary, "11N", 26911); if (!is.null(p)) st_transform(p, st_crs(ownership)) },
  { p <- make_points(hole_summary, "12N", 26912); if (!is.null(p)) st_transform(p, st_crs(ownership)) }
)
pts_v <- terra::vect(pts)
```

## Volume summary table

```{r tbl-volumes}
knitr::kable(
  volumes,
  col.names = c("Product", "Approved area (m³)", "Ownership area (m³)"),
  format.args = list(big.mark = ","),
  caption = paste0("Estimated peat volumes — ", params$project_id,
                   ". Harvestable values are net of the reclamation allowance."))
```

## Per-hole peat profiles by Von Post humification

```{r fig-profiles, fig.height=6}
# Non-peat layers carry no Von Post grade (band = NA); split them for the figure
# into forest floor (LFH / L-F-H folic) and mineral substrate (C/Cg, A/B).
forest_floor_pat <- "^(LFH|LF|FH|L|F|H)$"
grade_levels <- c("professional", "retail", "non_saleable", "forest_floor", "mineral")
band_cols <- c(professional = "#1a9850", retail = "#fee08b",
               non_saleable = "#8c510a", forest_floor = "#c8a165",
               mineral = "#9e9e9e")

p_prof <- intervals |>
  mutate(grade = dplyr::case_when(
           !is.na(band)                              ~ band,
           grepl(forest_floor_pat, toupper(trimws(horizon))) ~ "forest_floor",
           TRUE                                      ~ "mineral"),
         grade = factor(grade, levels = grade_levels)) |>
  ggplot(aes(x = plot_id,
             ymin = -depth_bot_cm, ymax = -depth_top_cm,
             colour = grade)) +
  geom_linerange(linewidth = 8) +
  scale_colour_manual(values = band_cols, name = "Layer", drop = FALSE,
                      labels = c(professional = "Professional (H1-H4)",
                                 retail = "Retail (H5-H6)",
                                 non_saleable = "Non-saleable (H7-H10)",
                                 forest_floor = "Forest floor (LFH)",
                                 mineral = "Mineral substrate (C)")) +
  scale_y_continuous(labels = ~ abs(.)) +
  labs(title = "Peat Profiles by Grade", x = "Test hole", y = "Depth (cm)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_prof
ggsave(here(params$output_dir, paste0(params$project_id, "_peat_profiles.png")),
       p_prof, width = 9, height = 5, dpi = 300)
```

## Per-hole peat profiles with the reclamation buffer

The same profiles with a **dashed red line** at the base of the harvestable
column (mineral depth − reclamation allowance). Everything below the line is the
deepest material left in place for reclamation. Provided as a separate figure so
it can be compared against the plain profiles above.

```{r fig-profiles-reclamation, fig.height=6}
reclaim_cm  <- round(stats::median(intervals$mineral_depth_cm - intervals$recl_line_cm))
plot_levels <- sort(unique(intervals$plot_id))
recl_label  <- paste0(reclaim_cm, " cm reclamation buffer")
recl_lines  <- intervals |>
  distinct(plot_id, recl_line_cm) |>
  mutate(xpos = match(plot_id, plot_levels),
         y    = -pmax(recl_line_cm, 0))   # clamp shallow holes to the surface

# Map the line to a linetype aesthetic so it gets its own legend entry (drawn
# red via override.aes), rather than explaining it in a subtitle.
p_prof_recl <- p_prof +
  geom_segment(data = recl_lines, inherit.aes = FALSE,
               aes(x = xpos - 0.45, xend = xpos + 0.45, y = y, yend = y,
                   linetype = recl_label),
               colour = "red", linewidth = 0.7) +
  scale_linetype_manual(name = NULL, values = setNames("dashed", recl_label),
                        guide = guide_legend(override.aes = list(colour = "red"))) +
  labs(title = "Peat Profiles by Grade with Reclamation Buffer")

p_prof_recl
ggsave(here(params$output_dir, paste0(params$project_id, "_peat_profiles_reclamation.png")),
       p_prof_recl, width = 9, height = 5, dpi = 300)
```

## Cross-section (stratigraphic fence of logged test holes)

A vertical section along a chosen transect of test holes (`params$section_holes`).
Each hole is drawn as its **real logged grade column** at its true along-transect
distance; the surface (0) and the mineral base are connected between holes. The
ground between holes is left blank — this is honest logged stratigraphy, not
depth-resolved interpolation (Von Post flips with depth, so grade-at-depth is not
interpolated; see DESIGN).

```{r fig-cross-section, fig.height=5}
sec_ids <- trimws(strsplit(params$section_holes, ",")[[1]])
sec_ids <- sec_ids[sec_ids != ""]

# Chosen holes in transect order, with coordinates and total depth.
sec_xy <- hole_summary |>
  distinct(plot_id, gps_easting, gps_northing, mineral_depth_cm)
sec_xy <- sec_xy[match(sec_ids, sec_xy$plot_id), ]
if (anyNA(sec_xy$gps_easting))
  stop("section_holes not found in data: ",
       paste(sec_ids[is.na(sec_xy$gps_easting)], collapse = ", "))

# Cumulative straight-line distance along the polyline through the holes (m).
sec_xy$dist_m <- cumsum(c(0, sqrt(diff(sec_xy$gps_easting)^2 +
                                  diff(sec_xy$gps_northing)^2)))

# Real logged intervals for those holes, classified to the same grade palette.
sec_intervals <- intervals |>
  filter(plot_id %in% sec_ids) |>
  mutate(grade = dplyr::case_when(
           !is.na(band)                                      ~ band,
           grepl(forest_floor_pat, toupper(trimws(horizon))) ~ "forest_floor",
           TRUE                                              ~ "mineral"),
         grade = factor(grade, levels = grade_levels)) |>
  left_join(sec_xy[, c("plot_id", "dist_m")], by = "plot_id")

bar_w <- 0.012 * max(sec_xy$dist_m)   # core column half-width (m)

p_section <- ggplot() +
  geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.4) +          # surface
  geom_line(data = sec_xy, aes(x = dist_m, y = -mineral_depth_cm),          # mineral base
            colour = "grey40", linetype = "dashed") +
  geom_rect(data = sec_intervals,                                           # logged cores
            aes(xmin = dist_m - bar_w, xmax = dist_m + bar_w,
                ymin = -depth_bot_cm, ymax = -depth_top_cm, fill = grade)) +
  geom_text(data = sec_xy, aes(x = dist_m, y = 20, label = plot_id),
            angle = 90, hjust = 0, size = 2.6) +
  scale_fill_manual(values = band_cols, name = "Layer", drop = FALSE,
                    labels = c(professional = "Professional (H1-H4)",
                               retail = "Retail (H5-H6)",
                               non_saleable = "Non-saleable (H7-H10)",
                               forest_floor = "Forest floor (LFH)",
                               mineral = "Mineral substrate (C)")) +
  scale_y_continuous(labels = ~ abs(.), expand = expansion(mult = c(0.04, 0.22))) +
  labs(title = "Peat Cross-Section (logged test holes along transect)",
       subtitle = paste(sec_ids, collapse = " → "),
       x = "Distance along transect (m)", y = "Depth (cm)") +
  theme_bw()

p_section
ggsave(here(params$output_dir, paste0(params$project_id, "_peat_cross_section.png")),
       p_section, width = 10, height = 5, dpi = 300)
```

## Harvestable thickness per hole

```{r fig-thickness}
thick_long <- hole_summary |>
  select(plot_id, Professional = prof_harv_cm, Retail = retail_harv_cm) |>
  tidyr::pivot_longer(-plot_id, names_to = "grade", values_to = "thickness_cm")

p_thick <- thick_long |>
  ggplot(aes(x = reorder(plot_id, -thickness_cm), y = thickness_cm, fill = grade)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(Professional = "#1a9850", Retail = "#fee08b")) +
  labs(title = "Harvestable Peat Thickness by Hole (net of reclamation)",
       x = "Test hole", y = "Thickness (cm)", fill = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_thick
ggsave(here(params$output_dir, paste0(params$project_id, "_peat_thickness.png")),
       p_thick, width = 9, height = 5, dpi = 300)
```

## Interpolated thickness maps with boundaries

```{r fig-maps, fig.height=6}
# Read back the surfaces written in 03_transform.
surf_path <- function(nm)
  here(params$output_dir, paste0(params$project_id, "_thickness_", nm, ".tif"))

plot_surface <- function(nm, title, contour = FALSE) {
  r <- terra::rast(surf_path(nm))
  r <- terra::mask(r, terra::vect(ownership))           # clip to study area

  # Contour levels are derived from the raster's own pretty breaks, so the
  # interval always matches the colour-scale legend (~0.5 m for professional,
  # ~0.1 m for retail, etc.) without hard-coding — adapts to any future surface.
  # The actual interval is reflected in the title.
  if (contour) {
    lev  <- pretty(range(terra::values(r), na.rm = TRUE))
    lev  <- lev[lev > 0]
    step <- if (length(lev) > 1) signif(lev[2] - lev[1], 3) else NA
    title <- paste0(title, " - ", step, " m contours")
  }

  # Optional Esri World Imagery backdrop (licensed for reports with attribution).
  if (isTRUE(params$basemap) && requireNamespace("maptiles", quietly = TRUE)) {
    bm <- maptiles::get_tiles(terra::ext(r), provider = "Esri.WorldImagery",
                              crop = TRUE, project = TRUE)
    terra::plotRGB(bm); terra::plot(r, alpha = 0.6, add = TRUE)
  } else {
    terra::plot(r, main = title, xlab = "Easting (m)", ylab = "Northing (m)",
                colNA = "grey85",                       # area outside the study boundary
                plg = list(title = "Peat Depth (m)", title.cex = 0.8))
  }
  terra::plot(terra::vect(ownership), border = "grey20", lwd = 1, add = TRUE)
  terra::plot(terra::vect(approved),  border = "red",    lwd = 2, add = TRUE)

  if (contour) {
    # White lines and labels read well across the viridis range (low dark to high
    # bright); only the brightest yellow caps are marginal, and the haloed
    # plot_id point labels remain the primary reference markers.
    terra::contour(r, add = TRUE, levels = lev, col = "white",
                   lwd = 0.8, labcex = 0.7, drawlabels = TRUE)
  }

  # Test holes, labelled with plot_id so the map ties back to the profile chart.
  terra::plot(pts_v, add = TRUE, col = "black", pch = 21, bg = "white", cex = 1.1)
  terra::text(pts_v, labels = "plot_id", pos = 3, cex = 0.55,
              halo = TRUE, hc = "white", hw = 0.2)

  # Boundary legend (the red / grey outlines on the map). Use terra::add_legend,
  # which targets the map panel — base legend() does not survive terra's
  # colour-bar layout and renders into the wrong region.
  terra::add_legend("bottomleft",
    legend = c("Reno Exploration Lands", "Tactful Developments Ltd. Boundary"),
    col = c("red", "grey20"), lwd = c(2, 1), lty = 1,
    bg = "white", box.col = "grey70", cex = 0.7)
}

save_map <- function(nm, title, suffix = "", contour = FALSE) {
  png(here(params$output_dir,
           paste0(params$project_id, "_map_", nm, suffix, ".png")),
      width = 1800, height = 1500, res = 200)
  plot_surface(nm, title, contour = contour)
  dev.off()
}

# Plain thickness maps (now with Easting/Northing axes and a legend title)
save_map("professional", "Professional peat thickness (m)")
save_map("retail",       "Retail peat thickness (m)")

# Second set: the same surfaces with contour lines (interval auto-set per surface)
save_map("professional", "Professional peat thickness (m)",
         suffix = "_contour", contour = TRUE)
save_map("retail",       "Retail peat thickness (m)",
         suffix = "_contour", contour = TRUE)

# Show inline as well
plot_surface("professional", "Professional peat thickness (m)")
plot_surface("retail",       "Retail peat thickness (m)")
plot_surface("professional", "Professional peat thickness (m)", contour = TRUE)
plot_surface("retail",       "Retail peat thickness (m)", contour = TRUE)
```
