---
title: "Peat Volume Uncertainty"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  ownership_boundary: "data-raw/projects/ownership_boundary.shp"
  waterbody_setback_m: 45  # for the table note — must match 03_transform
  utility_buffer_m: 30     # for the table note — must match 03_transform
  n_boot: 1000             # bootstrap resamples (0 to skip); ~2 min at 1000 on a 25-hole grid
  boot_ci: 0.90            # central interval width for the bootstrap range
  power_range: !r c(1, 2, 3)  # IDW power values for the parameter-sensitivity span
  jitter_m: 0.5            # coordinate jitter (m) to de-duplicate bootstrap-resampled cores
  seed: 42
  output_dir: "reports/figures"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(sf)
library(terra)
library(here)
library(ggplot2)
set.seed(params$seed)
```

This step quantifies **interpolation and sampling** uncertainty around the base
IDW volumes. It produces three *distinct* kinds of range — they are reported
separately and are **not additive**:

  - **Bootstrap interval** (statistical): resample the test holes, re-interpolate
    and re-integrate, take central percentiles. Captures the dominant *sampling*
    uncertainty from having only ~25 holes. This is the client-facing range.
  - **Parameter sensitivity** (scenario): volume under alternative but defensible
    IDW power values — a deterministic min–max, not a probability. Appendix.
  - **LOOCV RMSE** (validation): leave-one-out prediction error of the surface at
    withheld holes — a goodness-of-fit metric, not a volume range. Appendix.

It does **not** quantify Von Post / measurement / banding error.

```{r load}
hole_summary <- readRDS(here("data-raw", paste0(params$project_id, "_peat_hole_summary.rds")))
geom         <- readRDS(here("data-raw", paste0(params$project_id, "_peat_geom.rds")))
base_vol     <- readRDS(here("data-raw", paste0(params$project_id, "_peat_harvestable.rds")))

ownership <- st_read(here(params$ownership_boundary), quiet = TRUE)

# Interpolation points in the analysis (ownership) CRS — same construction as 03.
make_points <- function(df, zone, epsg) {
  sub <- dplyr::filter(df, utm_zone == zone)
  if (nrow(sub) == 0) return(NULL)
  st_as_sf(sub, coords = c("gps_easting", "gps_northing"), crs = epsg, remove = FALSE)
}
pts <- dplyr::bind_rows(
  { p <- make_points(hole_summary, "11N", 26911); if (!is.null(p)) st_transform(p, st_crs(ownership)) },
  { p <- make_points(hole_summary, "12N", 26912); if (!is.null(p)) st_transform(p, st_crs(ownership)) }
)

# Reuse the exact grid + clip geometry from 03_transform.
template  <- terra::rast(terra::vect(ownership), resolution = geom$res_m)
radius    <- geom$radius
power0    <- geom$idw_power
cell_area <- prod(terra::res(template))
cov_app <- terra::rasterize(terra::vect(sf::st_sf(geometry = geom$approved_harv)),  template, cover = TRUE)
cov_own <- terra::rasterize(terra::vect(sf::st_sf(geometry = geom$ownership_harv)), template, cover = TRUE)

# Core table: coordinates + the two commercial band thicknesses (m).
cores <- data.frame(
  x      = sf::st_coordinates(pts)[, 1],
  y      = sf::st_coordinates(pts)[, 2],
  prof   = pts$prof_harv_m,
  retail = pts$retail_harv_m
)
```

```{r helpers}
band_label <- c(prof = "Professional", retail = "Retail")

# Interpolate a field from a core table over the grid, integrate over a coverage
# raster -> volume (m³). Coverage weighting (not a centre mask) keeps it
# consistent with 03 and resolution-robust.
interp_vol <- function(df, field, cov, power = power0) {
  v <- terra::vect(df, geom = c("x", "y"), crs = terra::crs(template))
  s <- terra::interpIDW(template, v, field = field, radius = radius, power = power)
  terra::global(s * cov, "sum", na.rm = TRUE)[1, 1] * cell_area
}
```

## Bootstrap interval (sampling uncertainty)

```{r bootstrap}
boot <- NULL
if (params$n_boot > 0) {
  combos <- list(
    c(band = "prof",   area = "approved"),  c(band = "retail", area = "approved"),
    c(band = "prof",   area = "ownership"), c(band = "retail", area = "ownership")
  )
  draws <- matrix(NA_real_, nrow = params$n_boot, ncol = length(combos))
  for (b in seq_len(params$n_boot)) {
    d <- cores[sample(nrow(cores), replace = TRUE), ]
    # Jitter de-duplicates coincident resampled cores (IDW can't take 1/d=Inf);
    # 0.5 m is negligible on a ~25 m grid.
    d$x <- d$x + rnorm(nrow(d), 0, params$jitter_m)
    d$y <- d$y + rnorm(nrow(d), 0, params$jitter_m)
    for (k in seq_along(combos)) {
      cov_k <- if (combos[[k]]["area"] == "approved") cov_app else cov_own
      draws[b, k] <- interp_vol(d, combos[[k]]["band"], cov_k)
    }
  }
  lo <- (1 - params$boot_ci) / 2; hi <- 1 - lo
  boot <- data.frame(
    band = sapply(combos, `[`, "band"),
    area = sapply(combos, `[`, "area"),
    p_lo = apply(draws, 2, quantile, lo, na.rm = TRUE),
    p_hi = apply(draws, 2, quantile, hi, na.rm = TRUE)
  )
}
```

```{r tbl-headline}
# CLIENT-FACING REPORT TABLE: full net-harvestable table (all rows) with a
# bootstrap interval on the two harvestable rows. Excluded rows show the point
# estimate (their uncertainty is secondary and would clutter the headline).
fmt <- function(x) formatC(round(x), big.mark = ",", format = "d")
band_key <- c("Professional peat — harvestable" = "prof",
              "Retail peat — harvestable"        = "retail")
ci <- function(product, area) {
  k <- band_key[product]
  if (is.null(boot) || is.na(k)) return("—")          # em-dash for excluded rows
  r <- boot[boot$band == k & boot$area == area, ]
  if (!nrow(r)) return("—")
  paste0(fmt(r$p_lo), " – ", fmt(r$p_hi))
}
headline <- data.frame(
  Product = base_vol$product,
  `Approved — estimate (m³)`     = fmt(base_vol$approved_m3),
  `Approved — 90% interval (m³)` = vapply(base_vol$product, ci, "", area = "approved"),
  `Ownership — estimate (m³)`    = fmt(base_vol$ownership_m3),
  `Ownership — 90% interval (m³)`= vapply(base_vol$product, ci, "", area = "ownership"),
  check.names = FALSE, row.names = NULL)
knitr::kable(headline, row.names = FALSE,
  caption = paste0("Net harvestable peat — ", params$project_id,
    ". Base IDW estimate with a ", round(params$boot_ci * 100), "% bootstrap interval (",
    params$n_boot, " resamples) on the harvestable rows, reflecting sampling ",
    "uncertainty from the limited number of test holes. Intervals are not applied ",
    "to the deduction rows."))
```

```{r note-headline, results='asis'}
# Built into a variable so it can be saved with the table and reused verbatim by
# the report (report_template.qmd), keeping the note traceable to this run's params.
note_headline <- sprintf("
*Notes on the table above:* The two harvestable rows plus the two deduction rows
(*Excluded — waterbody setback* and *Excluded — other*) sum to the commercial
in-place resource.

- The **90%% interval** is a bootstrap range: the test holes were resampled %d times, re-interpolated and re-integrated, and the central %d%% of results reported. It expresses **sampling uncertainty** from the limited number of holes — not measurement or Von Post error — and is a statistical range (distinct from the parameter-sensitivity span in the appendix, which is a deterministic scenario; the two are not combined).
- **Excluded — waterbody setback.** The harvestable peat that falls within the %d m waterbody setback — its marginal contribution, i.e. the volume that would return to the resource if that setback were reduced or removed. The setback is a %d m buffer applied around every waterbody, both polygons (lakes and ponds) and polylines (creeks and streams), across the entire study area **including the previously approved lands** — applied conservatively because any existing *Water Act* approval over the approved lands may have lapsed and would require updating under current conditions.
- **Excluded — other.** Harvestable peat removed for non-waterbody constraints — generally either the **%d m setback around low-pressure pipelines and underground utilities**, or the peat **beneath existing roads**, which is present but not harvestable because the road is constructed as fill/corduroy over intact peat. Where the expansion area applies, this row also includes the required property-boundary harvest setback.
",
  as.integer(params$n_boot), as.integer(params$boot_ci * 100),
  as.integer(params$waterbody_setback_m), as.integer(params$waterbody_setback_m),
  as.integer(params$utility_buffer_m))
cat(note_headline)
```

```{r fig-bootstrap, fig.height=5}
if (!is.null(boot)) {
  # Full, human-readable panel titles (e.g. "Approved Professional Peat").
  area_label <- c(approved = "Approved", ownership = "Ownership")
  band_label <- c(prof = "Professional", retail = "Retail")
  series_full <- function(area, band)
    paste(area_label[area], band_label[band], "Peat")

  dd <- as.data.frame(draws)
  names(dd) <- vapply(combos, function(c) series_full(c["area"], c["band"]), "")
  long <- tidyr::pivot_longer(dd, everything(),
                              names_to = "series", values_to = "volume_m3")

  # Per-panel markers: the reported point estimate (red line — the value in the
  # net-harvestable table) and the 90% bootstrap interval (shaded band + dashed
  # lines), so the distribution is read against the numbers in the table.
  base_product <- setNames(names(band_key), band_key)   # "prof" -> "Professional peat — harvestable"
  markers <- do.call(rbind, lapply(combos, function(c) {
    a <- c["area"]; b <- c["band"]
    est <- base_vol[[paste0(a, "_m3")]][base_vol$product == base_product[b]]
    r   <- boot[boot$band == b & boot$area == a, ]
    data.frame(series = series_full(a, b), est = est, p_lo = r$p_lo, p_hi = r$p_hi)
  }))

  g <- ggplot(long, aes(volume_m3)) +
    geom_rect(data = markers, inherit.aes = FALSE,
              aes(xmin = p_lo, xmax = p_hi, ymin = -Inf, ymax = Inf),
              fill = "red", alpha = 0.08) +
    geom_histogram(bins = 40, fill = "#2c7fb8", colour = "white") +
    geom_vline(data = markers, aes(xintercept = p_lo),
               colour = "red", linetype = "dashed", linewidth = 0.4) +
    geom_vline(data = markers, aes(xintercept = p_hi),
               colour = "red", linetype = "dashed", linewidth = 0.4) +
    geom_vline(data = markers, aes(xintercept = est),
               colour = "red", linewidth = 0.8) +
    facet_wrap(~series, scales = "free", ncol = 2) +
    scale_x_continuous(labels = scales::label_comma()) +
    labs(title = "Bootstrap distribution of harvestable volume",
         subtitle = paste0("Solid red line = reported estimate; shaded band and dashed lines = ",
                           round(params$boot_ci * 100), "% bootstrap interval (both from Table 2)."),
         x = "Volume (m³)", y = "Resamples") +
    theme_bw()
  print(g)
  ggsave(here(params$output_dir, paste0(params$project_id, "_peat_bootstrap.png")),
         g, width = 9, height = 5.5, dpi = 300)
}
```

## Appendix A — Model validation (LOOCV)

```{r loocv}
loocv_rmse_cm <- function(field) {
  obs  <- cores[[field]]
  pred <- vapply(seq_len(nrow(cores)), function(i) {
    v  <- terra::vect(cores[-i, ], geom = c("x", "y"), crs = terra::crs(template))
    si <- terra::interpIDW(template, v, field = field, radius = radius, power = power0)
    terra::extract(si, terra::vect(cores[i, ], geom = c("x", "y"),
                                   crs = terra::crs(template)))[1, 2]
  }, numeric(1))
  sqrt(mean((pred - obs)^2, na.rm = TRUE)) * 100   # m -> cm
}
loocv <- data.frame(
  Band = c("Professional (H1–H4)", "Retail (H5–H6)"),
  `LOOCV RMSE (cm)` = round(c(loocv_rmse_cm("prof"), loocv_rmse_cm("retail")), 1),
  check.names = FALSE)
knitr::kable(loocv, row.names = FALSE,
  caption = paste0("Leave-one-out cross-validation of the base IDW surface (power ",
                   power0, "): root-mean-square error of predicted vs observed ",
                   "harvestable thickness at each withheld test hole. A surface-",
                   "accuracy metric, not a volume range."))
```

## Appendix B — Parameter sensitivity (IDW power)

```{r sensitivity}
sens <- do.call(rbind, lapply(c("prof", "retail"), function(bd) {
  va <- sapply(params$power_range, function(pw) interp_vol(cores, bd, cov_app, pw))
  vo <- sapply(params$power_range, function(pw) interp_vol(cores, bd, cov_own, pw))
  data.frame(
    Band = band_label[bd],
    `Approved — base (m³)`  = fmt(va[params$power_range == power0]),
    `Approved — range (m³)` = paste0(fmt(min(va)), " – ", fmt(max(va))),
    `Ownership — base (m³)` = fmt(vo[params$power_range == power0]),
    `Ownership — range (m³)` = paste0(fmt(min(vo)), " – ", fmt(max(vo))),
    check.names = FALSE)
}))
knitr::kable(sens, row.names = FALSE,
  caption = paste0("Sensitivity of net harvestable volume to the IDW power ",
    "(values ", paste(params$power_range, collapse = ", "), "; base = ", power0,
    "). A deterministic scenario span under alternative interpolation settings, ",
    "not a statistical confidence interval."))
```

```{r save}
# headline + note are the report-ready net-harvestable table and its caption note;
# saved here so report_template.qmd consumes them directly instead of re-deriving
# the bootstrap-interval formatting.
saveRDS(list(bootstrap = boot, loocv = loocv, sensitivity = sens,
             n_boot = params$n_boot, boot_ci = params$boot_ci,
             headline = headline, note = note_headline),
        here("data-raw", paste0(params$project_id, "_peat_uncertainty.rds")))
write.csv(headline,
          here("data-raw", paste0(params$project_id, "_peat_harvestable_report.csv")),
          row.names = FALSE)
```
