---
title: "Peat Exploration & Resource Assessment"
subtitle: "`r params$project_name`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  docx:
    reference-doc: report_style.docx
    toc: true
    toc-depth: 3
    lof: true
    lot: true
    number-sections: true
  html:
    toc: true
    toc-location: left
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
params:
  project_id:   "PROJECT_ID"
  project_name: "Project Name"
  client:       "Client Name"
  prepared_by:  "Green Plan Ltd."
  figure_dir:   "reports/figures"
  map_style:    "contour"   # thickness-map style: "contour" or "plain"
  photo_manifest: ""        # CSV (columns: file, caption) for the Appendix C photo log
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(knitr)
library(here)

peat_field_csv <- here("data-raw", paste0(params$project_id, "_peat_field_summary.csv"))
peat_unc_rds   <- here("data-raw", paste0(params$project_id, "_peat_uncertainty.rds"))
has_peat       <- file.exists(peat_field_csv) && file.exists(peat_unc_rds)

# Read the pipeline outputs once here; every table/figure below references these
# objects so the headline volumes are guaranteed identical across the report.
peat_unc   <- if (has_peat) readRDS(peat_unc_rds) else NULL
peat_field <- if (has_peat) read.csv(peat_field_csv, check.names = FALSE,
                                     fileEncoding = "UTF-8") else NULL

# Project-prefixed figure path produced by analysis/peat/04_visualize.Rmd
figp <- function(name, ext = ".png") {
  here(params$figure_dir, paste0(params$project_id, "_", name, ext))
}

# Thickness-map style selector: "contour" uses the *_contour.png maps, "plain"
# uses the unsuffixed ones. Lets the report show one map per grade without
# editing chunks — flip params$map_style.
map_suffix <- if (identical(params$map_style, "plain")) "" else "_contour"
```

```{r}
#| label: missing-data-banner
#| output: asis
#| eval: !expr "!has_peat"
cat("> **No peat module outputs found for `", params$project_id,
    "`.** Run `analysis/peat/run_pipeline.R` (steps 01–05) to populate the ",
    "tables and figures in this report.\n", sep = "")
```

# Executive Summary

*[Summarise the harvestable peat resource: total area assessed, net harvestable
volume split by professional (H1–H4) and retail (H5–H6) grade, the bootstrap
uncertainty interval, and the volume remaining after harvest setbacks and
infrastructure exclusions are applied.]*

```{r}
#| label: tbl-exec-summary
#| tbl-cap: "Headline net harvestable peat volumes by product grade."
#| eval: !expr has_peat
knitr::kable(peat_unc$headline, row.names = FALSE)
```

# Introduction

## Background

*[Insert project background: who commissioned the exploration, the location and
size of the deposit, prior work, and the intended end use of the peat resource.]*

## Objectives

The objective of this exploration is to characterise the peat resource and
estimate the **net harvestable volume by product grade**. Peat is classified for
saleability using the Von Post humification scale, following the client product
specification:

- **Professional grade** — H1–H4 (low humification, fibric/mesic peat)
- **Retail grade** — H5–H6 (moderately decomposed peat)
- **Non-saleable** — H7–H10 and the basal reclamation allowance

## Site Description

*[Insert physical setting, wetland class/form, surface vegetation, hydrology, and
access constraints relevant to harvest planning.]*

# Methods

## Field Investigation

Test holes were advanced across the deposit on a survey grid. At each hole the
crew recorded AWCS wetland class, dominant surface bryophyte, total peat depth,
depth to water table, pH, electrical conductivity, and Von Post humification by
depth interval. Per-hole observations are tabulated in @tbl-peat-field (Appendix B).

## Product Grade Classification

Each depth interval was assigned to a product grade from its Von Post value using
the client specification above. Harvestable thickness by grade was accumulated
per hole, **net of a 0.5 m basal reclamation allowance** retained to support site
reclamation; this allowance is not counted as a saleable resource.

## Volume Estimation

Per-hole harvestable thicknesses were interpolated across the deposit by inverse
distance weighting (IDW) to produce continuous thickness surfaces for each
product grade. Volumes were computed by integrating the banded thickness surfaces
over the assessment area.

## Harvest Setbacks and Infrastructure Exclusions

Harvestable volumes were recomputed with regulatory and operational exclusions
applied — setbacks from watercourses and waterbodies, and buffers around roads
and other infrastructure — to give a constrained, operationally available volume.

## Uncertainty Analysis

Sampling uncertainty arising from the limited number of test holes was quantified
by bootstrap resampling of the holes. Leave-one-out cross-validation (LOOCV) and
sensitivity analysis on the interpolation parameters were used to check the
robustness of the volume estimates.

# Results

## Peat Thickness and Stratigraphy

Harvestable peat thickness at each test hole, split by product grade, is shown
in @fig-peat-thickness, and per-hole stratigraphy — with the 0.5 m basal
reclamation allowance marked — in @fig-peat-profiles-recl. Interpreted
cross-sections along selected test-hole transects are provided in Appendix A.

```{r}
#| label: fig-peat-thickness
#| fig-cap: "Harvestable peat thickness by test hole, split by product grade (net of the 0.5 m basal reclamation allowance)."
#| eval: !expr has_peat
knitr::include_graphics(figp("peat_thickness"))
```

```{r}
#| label: fig-peat-profiles-recl
#| fig-cap: "Per-hole peat profiles with the 0.5 m basal reclamation allowance removed."
#| eval: !expr has_peat
knitr::include_graphics(figp("peat_profiles_reclamation"))
```

## Harvestable Volume by Product Grade

Net harvestable volumes by grade, with a 90% bootstrap interval on the
harvestable rows, are given in @tbl-peat-harvestable. The professional- and
retail-grade thickness surfaces are mapped in @fig-peat-professional and
@fig-peat-retail.

```{r}
#| label: tbl-peat-harvestable
#| tbl-cap: "Net harvestable peat by product grade, with a 90% bootstrap interval (sampling uncertainty) on the harvestable rows."
#| eval: !expr has_peat
knitr::kable(peat_unc$headline, row.names = FALSE)
```

```{r}
#| label: note-peat-harvestable
#| output: asis
#| eval: !expr has_peat
cat(peat_unc$note)
```

```{r}
#| label: fig-peat-professional
#| fig-cap: "Professional-grade (H1–H4) harvestable peat thickness."
#| eval: !expr has_peat
knitr::include_graphics(figp(paste0("map_professional", map_suffix)))
```

```{r}
#| label: fig-peat-retail
#| fig-cap: "Retail-grade (H5–H6) harvestable peat thickness."
#| eval: !expr has_peat
knitr::include_graphics(figp(paste0("map_retail", map_suffix)))
```

## Setback-Constrained Volumes

With watercourse/waterbody setbacks and infrastructure buffers applied, the
operationally available footprint is reduced as shown in @fig-peat-professional-setbacks
and @fig-peat-retail-setbacks.

```{r}
#| label: fig-peat-professional-setbacks
#| fig-cap: "Professional-grade harvestable peat after harvest setbacks and infrastructure exclusions."
#| eval: !expr has_peat
knitr::include_graphics(figp("map_professional_setbacks"))
```

```{r}
#| label: fig-peat-retail-setbacks
#| fig-cap: "Retail-grade harvestable peat after harvest setbacks and infrastructure exclusions."
#| eval: !expr has_peat
knitr::include_graphics(figp("map_retail_setbacks"))
```

## Uncertainty

The bootstrap distribution of total harvestable volume (@fig-peat-bootstrap)
reflects sampling uncertainty from the limited number of test holes.

```{r}
#| label: fig-peat-bootstrap
#| fig-cap: "Bootstrap distribution of harvestable volume, reflecting sampling uncertainty from the limited number of test holes."
#| eval: !expr has_peat
knitr::include_graphics(figp("peat_bootstrap"))
```

# Conclusions

*[State the net harvestable resource by grade, the operationally available volume
after exclusions, the associated uncertainty, and any recommendations for further
investigation or harvest planning.]*

# Appendix A — Cross-Section Figures {.unnumbered}

Stratigraphic fence diagrams along selected test-hole transects. Each hole is
drawn as its real logged grade column at its true along-transect distance; the
mineral base, 0.5 m reclamation line, and water table are connected between
holes. The ground between holes is left blank — this is honest logged
stratigraphy, not depth-resolved interpolation. The overview map (@fig-xsec-overview)
locates every section; each section's test holes are labelled on the cores and in
its location inset.

Sections are lettered by location (north→south, then west→east) so they progress
sensibly across the location map, and the holes within each section are ordered
for left-to-right, top-to-bottom reading (west→east, then north→south) so each
section reads the same way as the map. Cross-sections are generated by
`analysis/peat/cross_sections.R`; the overview map is @fig-xsec-overview and the
lettered sections follow.

```{r}
#| label: fig-xsec-overview
#| fig-cap: "Cross-section location overview — test-hole transects A–I."
#| eval: !expr "file.exists(figp('cross_section_overview'))"
knitr::include_graphics(figp("cross_section_overview"))
```

```{r}
#| label: xsec-series
#| output: asis
#| eval: !expr has_peat
# Data-driven: emit every cross_section_<LETTER>.png present in figure_dir, in
# letter order, so the appendix matches however many sections cross_sections.R
# produced for this project (no fixed A–I list to keep in sync).
xfiles <- list.files(here(params$figure_dir),
                     pattern = "_cross_section_[A-Z]\\.png$", full.names = TRUE)
xfiles <- xfiles[startsWith(basename(xfiles), paste0(params$project_id, "_"))]
xfiles <- xfiles[order(basename(xfiles))]
for (f in xfiles) {
  letter <- sub(".*_cross_section_([A-Z])\\.png$", "\\1", basename(f))
  cat(sprintf("\n\n![Cross-Section %s.](%s)\n\n", letter, gsub("\\\\", "/", f)))
}
```

# Appendix B — Per-Hole Field Logs {.unnumbered}

Per-hole field measurements recorded at each test hole: AWCS wetland class,
dominant surface bryophyte, depth to mineral contact (peat depth), depth to water
table, pH, electrical conductivity, and harvestable thickness by product grade
(net of the 0.5 m basal reclamation allowance).

```{r}
#| label: tbl-peat-field
#| tbl-cap: "Per-hole field summary. Harvestable thicknesses are net of the 0.5 m basal reclamation allowance; \"—\" = not recorded (e.g. frozen ground) or not applicable (Upland anchor)."
#| eval: !expr has_peat
knitr::kable(peat_field)
```

# Appendix C — Photo Log {.unnumbered}

Site and test-hole photographs. To populate this log, add a CSV manifest with
`file` and `caption` columns — one row per photo, `file` paths relative to the
manifest's own folder — and pass its path to the report as the `photo_manifest`
parameter. Photos are numbered separately from the report figures.

```{r}
#| label: setup-photos
#| include: false
photo_csv <- if (nzchar(params$photo_manifest)) here(params$photo_manifest) else ""
photos    <- if (nzchar(photo_csv) && file.exists(photo_csv)) {
  read.csv(photo_csv, stringsAsFactors = FALSE, fileEncoding = "UTF-8")
} else {
  data.frame(file = character(), caption = character())
}
has_photos <- nrow(photos) > 0
```

```{r}
#| label: photos-placeholder
#| output: asis
#| eval: !expr "!has_photos"
cat("*[No photographs logged yet. Add photo files and rows to the project's",
    "`photos/photo_manifest.csv` (columns `file`, `caption`) to populate this",
    "appendix.]*")
```

```{r}
#| label: photo-log
#| output: asis
#| eval: !expr has_photos
base_dir <- dirname(photo_csv)
for (i in seq_len(nrow(photos))) {
  img <- photos$file[i]
  if (!file.exists(img)) img <- file.path(base_dir, img)   # resolve relative to manifest
  cat(sprintf("\n\n**Photo %d.** %s\n\n![](%s){width=80%%}\n\n",
              i, photos$caption[i], img))
}
```

---

*Report prepared by `r params$prepared_by` for `r params$client`.*
