---
title: "Peat Exploration & Resource Assessment"
subtitle: "`r params$project_name`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
bibliography: references.bib
link-citations: true
nocite: |
  @goa_awidd2015, @goa_wetlandpolicy2013, @goa_waterpermanence2015, @cssc1998,
  @terra2024, @rcoreteam2024
format:
  docx:
    reference-doc: report_style.docx
    toc: true
    toc-depth: 3
    lof: true
    lot: true
    number-sections: true
  html:
    toc: true
    toc-location: left
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
params:
  project_id:   "PROJECT_ID"
  project_name: "Project Name"
  client:       "Client Name"
  prepared_by:  "Green Plan Ltd."
  figure_dir:   "reports/figures"
  map_style:    "contour"   # thickness-map style: "contour" or "plain"
  photo_manifest: ""        # CSV (columns: file, caption) for the Appendix C photo log
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(knitr)
library(here)

peat_field_csv <- here("data-raw", paste0(params$project_id, "_peat_field_summary.csv"))
peat_unc_rds   <- here("data-raw", paste0(params$project_id, "_peat_uncertainty.rds"))
has_peat       <- file.exists(peat_field_csv) && file.exists(peat_unc_rds)

# Read the pipeline outputs once here; every table/figure below references these
# objects so the headline volumes are guaranteed identical across the report.
peat_unc   <- if (has_peat) readRDS(peat_unc_rds) else NULL
peat_field <- if (has_peat) read.csv(peat_field_csv, check.names = FALSE,
                                     fileEncoding = "UTF-8") else NULL

# Project-prefixed figure path produced by analysis/peat/04_visualize.Rmd
figp <- function(name, ext = ".png") {
  here(params$figure_dir, paste0(params$project_id, "_", name, ext))
}

# Thickness-map style selector: "contour" uses the *_contour.png maps, "plain"
# uses the unsuffixed ones. Lets the report show one map per grade without
# editing chunks — flip params$map_style.
map_suffix <- if (identical(params$map_style, "plain")) "" else "_contour"
```

```{r}
#| label: missing-data-banner
#| output: asis
#| eval: !expr "!has_peat"
cat("> **No peat module outputs found for `", params$project_id,
    "`.** Run `analysis/peat/run_pipeline.R` (steps 01–05) to populate the ",
    "tables and figures in this report.\n", sep = "")
```

# Executive Summary

*[Summarise the harvestable peat resource: total area assessed, net harvestable
volume split by professional (H1–H4) and retail (H5–H6) grade, the bootstrap
uncertainty interval, and the volume remaining after harvest setbacks and
infrastructure exclusions are applied.]*

```{r}
#| label: tbl-exec-summary
#| tbl-cap: "Headline net harvestable peat volumes by product grade."
#| eval: !expr has_peat
knitr::kable(peat_unc$headline, row.names = FALSE)
```

# Introduction

## Background

*[Insert project background: who commissioned the exploration, the location and
size of the deposit, prior work, and the intended end use of the peat resource.]*

## Objectives

The objective of this exploration is to characterise the peat resource and
estimate the **net harvestable volume by product grade**. Peat is classified for
saleability using the Von Post humification scale [@vonpost1922; @stanek1977],
following the client product
specification:

- **Professional grade** — H1–H4 (low humification, fibric/mesic peat)
- **Retail grade** — H5–H6 (moderately decomposed peat)
- **Non-saleable** — H7–H10 and the basal reclamation allowance

## Site Description

*[Insert physical setting, wetland class/form, surface vegetation, hydrology, and
access constraints relevant to harvest planning.]*

The study area is shown in @fig-area-overview — a drone orthomosaic with the
ownership and approved-harvest boundaries and the test-hole network overlaid.

```{r}
#| label: fig-area-overview
#| fig-cap: "Area overview — drone orthomosaic with ownership and approved-harvest boundaries and labelled test holes."
#| eval: !expr "file.exists(figp('area_overview'))"
knitr::include_graphics(figp("area_overview"))
```

# Methods

## Field Investigation

Test holes were advanced across the deposit on a survey grid. At each hole the
crew recorded AWCS wetland class [@goa_awcs2015], dominant surface bryophyte, total peat depth,
depth to water table, pH, electrical conductivity, and Von Post humification by
depth interval. Per-hole observations are tabulated in @tbl-peat-field (Appendix B).

## Product Grade Classification

Each depth interval was assigned to a product grade from its Von Post value using
the client specification above. Harvestable thickness by grade was accumulated
per hole, **net of a 0.5 m basal reclamation allowance** retained to support site
reclamation; this allowance is not counted as a saleable resource.

## Volume Estimation

Per-hole harvestable thicknesses were interpolated across the deposit by inverse
distance weighting (IDW) to produce continuous thickness surfaces for each
product grade. Volumes were computed by integrating the banded thickness surfaces
over the assessment area.

## Harvest Setbacks and Infrastructure Exclusions

Harvestable volumes were recomputed with regulatory and operational exclusions
applied — setbacks from watercourses and waterbodies, and buffers around roads
and other infrastructure — to give a constrained, operationally available volume.

## Uncertainty Analysis

Sampling uncertainty arising from the limited number of test holes was quantified
by bootstrap resampling of the holes. Leave-one-out cross-validation (LOOCV) and
sensitivity analysis on the interpolation parameters were used to check the
robustness of the volume estimates.

# Results

## Peat Thickness and Stratigraphy

Harvestable peat thickness at each test hole, split by product grade, is shown
in @fig-peat-thickness, and per-hole stratigraphy — with the 0.5 m basal
reclamation allowance marked — in @fig-peat-profiles-recl. Interpreted
cross-sections along selected test-hole transects are provided in Appendix A.

```{r}
#| label: fig-peat-thickness
#| fig-cap: "Harvestable peat thickness by test hole, split by product grade (net of the 0.5 m basal reclamation allowance)."
#| eval: !expr has_peat
knitr::include_graphics(figp("peat_thickness"))
```

```{r}
#| label: fig-peat-profiles-recl
#| fig-cap: "Per-hole peat profiles with the 0.5 m basal reclamation allowance removed."
#| eval: !expr has_peat
knitr::include_graphics(figp("peat_profiles_reclamation"))
```

## Harvestable Volume by Product Grade

Net harvestable volumes by grade, with a 90% bootstrap interval on the
harvestable rows, are given in @tbl-peat-harvestable. The professional- and
retail-grade thickness surfaces are mapped in @fig-peat-professional and
@fig-peat-retail.

```{r}
#| label: tbl-peat-harvestable
#| tbl-cap: "Net harvestable peat by product grade, with a 90% bootstrap interval (sampling uncertainty) on the harvestable rows."
#| eval: !expr has_peat
knitr::kable(peat_unc$headline, row.names = FALSE)
```

```{r}
#| label: note-peat-harvestable
#| output: asis
#| eval: !expr has_peat
# Render the note as a styled table footnote (small italic, via the Table Note
# paragraph style in report_style.docx) directly beneath the table.
cat("\n::: {custom-style=\"Table Note\"}\n", peat_unc$note, "\n:::\n", sep = "")
```

```{r}
#| label: fig-peat-professional
#| fig-cap: "Professional-grade (H1–H4) harvestable peat thickness."
#| eval: !expr has_peat
knitr::include_graphics(figp(paste0("map_professional", map_suffix)))
```

```{r}
#| label: fig-peat-retail
#| fig-cap: "Retail-grade (H5–H6) harvestable peat thickness."
#| eval: !expr has_peat
knitr::include_graphics(figp(paste0("map_retail", map_suffix)))
```

## Setback-Constrained Volumes

With watercourse/waterbody setbacks and infrastructure buffers applied, the
operationally available footprint is reduced as shown in @fig-peat-professional-setbacks
and @fig-peat-retail-setbacks.

```{r}
#| label: fig-peat-professional-setbacks
#| fig-cap: "Professional-grade harvestable peat after harvest setbacks and infrastructure exclusions."
#| eval: !expr has_peat
knitr::include_graphics(figp("map_professional_setbacks"))
```

```{r}
#| label: fig-peat-retail-setbacks
#| fig-cap: "Retail-grade harvestable peat after harvest setbacks and infrastructure exclusions."
#| eval: !expr has_peat
knitr::include_graphics(figp("map_retail_setbacks"))
```

## Uncertainty

The volume estimates depend on two things the reader should be able to weigh:
how the thickness surface is interpolated between test holes, and how much the
estimate could move given that only a limited number of holes were drilled. Both
are made explicit below.

### Interpolation method and assumptions

Harvestable thickness is interpolated between test holes by **inverse distance
weighting (IDW)** [@shepard1968] — each grid cell is a distance-weighted average of the
surrounding holes, so nearer holes carry more weight. The settings are:

- **Distance-decay power = 2** (the standard IDW exponent): weight falls off with
  the square of distance, a moderate setting that neither over-smooths nor lets a
  single nearby hole dominate.
- **Global search** (all holes inform every cell), appropriate for the hole count
  and spacing here; no maximum search radius is imposed.
- **Grid resolution set from the data** — the mean nearest-neighbour spacing
  between holes divided by ten (clamped to 5–25 m). Resolution governs how
  finely the boundaries and exclusions are clipped when integrating volume; it
  does not change the interpolated values, which are fixed by hole density.
- **Volume by fractional cell coverage** — each cell contributes its thickness
  times the fraction of its area inside the boundary (and outside any exclusion),
  so setbacks and corridors narrower than a cell are captured proportionally
  rather than rounded in or out.

IDW is an exact interpolator (it honours the logged value at each hole) and makes
no distributional assumptions about the peat. Its main limitation is the one
inherent to any interpolation from sparse points: between widely spaced holes the
surface is a smooth average and will not capture unobserved local variation. The
sensitivity of the estimate to the IDW power is reported in the parameter-
sensitivity analysis (a deterministic scenario span, separate from the bootstrap
range below).

### Sampling uncertainty (bootstrap)

The dominant uncertainty is **sampling uncertainty** — the deposit was
characterised from a limited number of test holes, and a different set of holes
would have produced a somewhat different surface. This is quantified by
**bootstrap resampling** [@efron1993]: the holes are resampled with replacement, the thickness
surface is re-interpolated and the volume re-integrated, and this is repeated many
times to build a distribution of plausible volumes. The central 90% of that
distribution is the interval reported in @tbl-peat-harvestable.

@fig-peat-bootstrap shows the resulting distribution for each grade and area. The
**solid red line is the reported estimate** (the base IDW value in
@tbl-peat-harvestable) and the **shaded band the 90% interval**, so the spread
around the headline number is visible directly. The bootstrap reflects sampling
uncertainty only — it does not include measurement or Von Post-classification
error, which are treated as secondary here.

```{r}
#| label: fig-peat-bootstrap
#| fig-cap: "Bootstrap distribution of harvestable volume by grade and area. Solid red line = reported estimate; shaded band and dashed lines = 90% bootstrap interval (both from the net-harvestable table)."
#| eval: !expr has_peat
knitr::include_graphics(figp("peat_bootstrap"))
```

# Conclusions

*[State the net harvestable resource by grade, the operationally available volume
after exclusions, the associated uncertainty, and any recommendations for further
investigation or harvest planning.]*

# Closure

This report is valid for three years past the field assessment date to ensure
that the peat resource characterization remains representative of site
conditions. Green Plan advises that any necessary approvals, licenses, and
permits are secured prior to the commencement of any harvest or construction
activities. All regulatory and mitigation requirements should be reviewed prior
to further activities to ensure that they are understood and implemented as
described.

This Peat Exploration & Resource Assessment was prepared to characterize the peat
resource within the vicinity of the Project, estimate the net harvestable volume
by product grade, and to support planning and applications for the provincial and
federal approvals required for the proposed peat harvest. This report presents
information available at the time of writing, as well as best management practices
and provincial and federal regulations effective at the time of writing. If you
have any questions or require additional details, please contact the undersigned.

Sincerely,

+-----------------------------------------------------------+
| **Report prepared by:**                                   |
+-----------------------------------------------------------+
|                                                           |
|                                                           |
+-----------------------------------------------------------+
| *[Name], [Designation]*                                   |
|                                                           |
| *[Title], Green Plan Ltd.*                                |
+-----------------------------------------------------------+
| **Report reviewed by:**                                   |
+-----------------------------------------------------------+
|                                                           |
|                                                           |
+-----------------------------------------------------------+
| *[Name], [Designation]*                                   |
|                                                           |
| *[Title], Green Plan Ltd.*                                |
+-----------------------------------------------------------+

# References {.unnumbered}

::: {#refs}
:::

# Appendix A — Record Searches {.unnumbered}

*[Insert the results of regulatory and historical record searches — e.g. Alberta
Environment and Protected Areas approvals and Water Act records, historical land
use and air photos, and any prior peat or wetland investigations relevant to the
deposit.]*

# Appendix B — Per-Hole Field Logs {.unnumbered}

Per-hole field measurements recorded at each test hole: AWCS wetland class,
dominant surface bryophyte, depth to mineral contact, depth to water table, pH,
electrical conductivity, and harvestable thickness by product grade (net of the
0.5 m basal reclamation allowance).

```{r}
#| label: tbl-peat-field
#| tbl-cap: "Per-hole field summary. Harvestable thicknesses are net of the 0.5 m basal reclamation allowance; \"—\" = not recorded (e.g. frozen ground) or not applicable (Upland anchor)."
#| eval: !expr has_peat
knitr::kable(peat_field, row.names = FALSE)
```

# Appendix C — Cross-Section Figures {.unnumbered}

Stratigraphic fence diagrams along selected test-hole transects. Each hole is
drawn as its real logged grade column at its true along-transect distance; the
mineral base, 0.5 m reclamation line, and water table are connected between
holes. The ground between holes is left blank — this is honest logged
stratigraphy, not depth-resolved interpolation. The overview map (@fig-xsec-overview)
locates every section; each section's test holes are labelled on the cores and in
its location inset.

Sections are lettered by location (north→south, then west→east) so they progress
sensibly across the location map, and the holes within each section are ordered
for left-to-right, top-to-bottom reading (west→east, then north→south) so each
section reads the same way as the map. Cross-sections are generated by
`analysis/peat/cross_sections.R`; the overview map is @fig-xsec-overview and the
lettered sections follow.

```{r}
#| label: fig-xsec-overview
#| fig-cap: "Cross-section location overview — test-hole transects A–I."
#| eval: !expr "file.exists(figp('cross_section_overview'))"
knitr::include_graphics(figp("cross_section_overview"))
```

```{r}
#| label: xsec-series
#| output: asis
#| eval: !expr has_peat
# Data-driven: emit every cross_section_<LETTER>.png present in figure_dir, in
# letter order, so the appendix matches however many sections cross_sections.R
# produced for this project (no fixed A–I list to keep in sync).
xfiles <- list.files(here(params$figure_dir),
                     pattern = "_cross_section_[A-Z]\\.png$", full.names = TRUE)
xfiles <- xfiles[startsWith(basename(xfiles), paste0(params$project_id, "_"))]
xfiles <- xfiles[order(basename(xfiles))]
for (f in xfiles) {
  letter <- sub(".*_cross_section_([A-Z])\\.png$", "\\1", basename(f))
  # Fixed width (landscape page) so the sections are a consistent size rather
  # than auto-scaled; keepNext (added by the finaliser) holds the caption with it.
  cat(sprintf("\n\n![Cross-Section %s.](%s){width=9in}\n\n", letter, gsub("\\\\", "/", f)))
}
```

# Appendix D — Photo Log {.unnumbered}

Site and test-hole photographs. To populate this log, add a CSV manifest with
`file` and `caption` columns — one row per photo, `file` paths relative to the
manifest's own folder — and pass its path to the report as the `photo_manifest`
parameter. Photos are numbered separately from the report figures.

```{r}
#| label: setup-photos
#| include: false
photo_csv <- if (nzchar(params$photo_manifest)) here(params$photo_manifest) else ""
photos    <- if (nzchar(photo_csv) && file.exists(photo_csv)) {
  read.csv(photo_csv, stringsAsFactors = FALSE, fileEncoding = "UTF-8")
} else {
  data.frame(file = character(), caption = character())
}
has_photos <- nrow(photos) > 0
```

```{r}
#| label: photos-placeholder
#| output: asis
#| eval: !expr "!has_photos"
cat("*[No photographs logged yet. Add photo files and rows to the project's",
    "`photos/photo_manifest.csv` (columns `file`, `caption`) to populate this",
    "appendix.]*")
```

```{r}
#| label: photo-log
#| output: asis
#| eval: !expr has_photos
base_dir <- dirname(photo_csv)
for (i in seq_len(nrow(photos))) {
  img <- photos$file[i]
  if (!file.exists(img)) img <- file.path(base_dir, img)   # resolve relative to manifest
  cat(sprintf("\n\n**Photo %d.** %s\n\n![](%s){width=80%%}\n\n",
              i, photos$caption[i], img))
}
```

---

*Report prepared by `r params$prepared_by` for `r params$client`.*
