---
title: "Wetland Assessment & Impact Report (WAIR)"
subtitle: "`r params$project_name`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  docx:
    reference-doc: report_style.docx
    toc: true
    toc-depth: 3
    number-sections: true
  html:
    toc: true
    toc-location: left
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
params:
  project_id:   "PROJECT_ID"
  project_name: "Project Name"
  client:       "Client Name"
  prepared_by:  "Green Plan Ltd."
  figure_dir:   "reports/figures"
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(knitr)
library(here)

# Show a figure if it has been generated, otherwise emit a pending placeholder
# so the template renders before the vegetation/soils/water pipelines have run.
show_fig <- function(file, what) {
  p <- here(params$figure_dir, file)
  if (file.exists(p)) {
    knitr::include_graphics(p)
  } else {
    knitr::asis_output(paste0("*[Figure pending: ", what, "]*"))
  }
}
```

# Executive Summary

*[Summarise the wetlands identified, their AWCS classification, the WAIR
significance scores, the proposed development, and the assessed impacts and
recommended mitigation.]*

# Introduction

## Project Background

*[Insert project background and the purpose of the assessment.]*

## Proposed Development

*[Describe the proposed development, its footprint, and its spatial relationship
to the wetlands assessed.]*

## Regulatory Framework

This assessment is conducted under the **Alberta Wetland Policy** and aligns with:

- **Alberta Wetland Classification System (AWCS)** — wetland class/form/type
- **Alberta Wetland Identification and Delineation Directive (AWIDD)** — delineation
- **Wetland Assessment and Impact Rating (WAIR)** — significance scoring
- **Guidance to Determining Water Permanence** — permanence assessment

# Methods

## Wetland Delineation

Wetland boundaries were delineated following AWIDD, using plot-based vegetation,
soils, and hydrology indicators at paired wetland/upland plots.

## Vegetation Assessment

Vascular plant cover was recorded by plot and each species assigned a wetland
indicator status (UPL, FACU, FAC, FACW, OBL). Plot-level UPL–OBL summaries and
ANPC wetland plant community affinities were derived for delineation and scoring.

## Soils Assessment

Soil profiles were described and classified following the Canadian System of Soil
Classification (CSSC). Hydric soil indicators (low chroma matrices, redoximorphic
features) were identified from Munsell notation; peat was characterised on the
Von Post humification scale.

## Water Permanence

Water permanence was assessed from multi-date satellite imagery using NDWI pixel
classification, supplemented by LiDAR-derived depression analysis, and assigned
to the Alberta permanence typology.

# Wetland Inventory and Classification

*[Insert the inventory of delineated wetlands with AWCS class, form, and type, and
a location figure.]*

```{r}
#| label: fig-wetland-inventory
#| fig-cap: "Delineated wetlands and assessment plots."
show_fig("fig_wetland_inventory.png", "wetland inventory map")
```

# Assessment Results

## Vegetation

Wetland indicator cover by plot is shown in @fig-upl-obl, native versus invasive
cover in @fig-invasive, and noxious weed occurrence in @fig-noxious.

```{r}
#| label: fig-upl-obl
#| fig-cap: "Wetland indicator cover by plot."
show_fig("fig_upl_obl_by_plot.png", "wetland indicator cover by plot")
```

```{r}
#| label: fig-invasive
#| fig-cap: "Native vs invasive cover by plot."
show_fig("fig_native_invasive_by_plot.png", "native vs invasive cover by plot")
```

```{r}
#| label: fig-noxious
#| fig-cap: "Noxious weed species heat map."
show_fig("fig_noxious_heatmap.png", "noxious weed heat map")
```

## Soils

Peat profiles classified by Von Post humification are presented in @fig-soils-vonpost.

```{r}
#| label: fig-soils-vonpost
#| fig-cap: "Peat profiles by Von Post humification."
show_fig("fig_peat_vonpost.png", "Von Post peat profiles")
```

## Water Permanence

*[Insert water permanence figures and interpretation.]*

# WAIR Scoring

The WAIR significance score combines vegetation, soils, water permanence, and
landscape-context inputs into a composite rating used in the impact assessment.

*[Insert the WAIR scoring table and the derived significance class per wetland.]*

# Impact Assessment

## Direct and Indirect Impacts

*[Assess direct (footprint) and indirect (hydrological, edge) impacts of the
proposed development on each wetland, referenced to its WAIR significance.]*

## Mitigation and Compensation

*[Describe avoidance, minimisation, and compensation measures consistent with the
Alberta Wetland Policy mitigation hierarchy.]*

# Conclusions and Recommendations

*[State the assessed significance, residual impacts after mitigation, and
recommendations.]*

---

*Report prepared by `r params$prepared_by` for `r params$client`.*
