---
title: "Wetland Monitoring Report"
subtitle: "`r params$project_name`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  docx:
    reference-doc: report_style.docx
    toc: true
    toc-depth: 3
    number-sections: true
  html:
    toc: true
    toc-location: left
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
params:
  project_id:    "PROJECT_ID"
  project_name:  "Project Name"
  client:        "Client Name"
  prepared_by:   "Green Plan Ltd."
  figure_dir:    "reports/figures"
  monitoring_year: ""
  baseline_year:   ""
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(knitr)
library(here)

# show_fig() — figure-if-exists-else-placeholder — shared with the WAIR
# template. figure_dir must be set before sourcing the helper.
figure_dir <- params$figure_dir
source(here("reports/_report_helpers.R"))
```

# Executive Summary

*[Summarise the monitoring event, condition relative to baseline and performance
targets, key trends, and whether the wetland is trending toward or away from its
target condition.]*

# Introduction

## Background and Monitoring Objectives

*[Insert the reason for monitoring (e.g. post-reclamation, compensation site,
ongoing condition tracking), the monitoring program objectives, and the
performance targets against which condition is assessed.]*

## Reference and Baseline Conditions

*[Describe the baseline condition (year `r params$baseline_year`) and any reference
sites used for comparison.]*

# Methods

## Monitoring Design

*[Describe the monitoring design: plot/transect network, sampling frequency, and
the monitoring events included in this report.]*

## Vegetation Monitoring

Vascular plant cover was recorded by plot and each species assigned a wetland
indicator status (UPL, FACU, FAC, FACW, OBL). Plot-level UPL–OBL summaries,
native/invasive composition, and noxious weed occurrence were derived for
comparison across monitoring events.

## Hydrology Monitoring

*[Describe hydrology monitoring: water-level/permanence measurements and the
imagery/LiDAR analysis used to characterise wetland hydrology over time.]*

## Soils Monitoring

Soil profiles were described following the CSSC; peat was characterised on the
Von Post humification scale where present.

# Results

## Vegetation Condition and Trend

Current-event wetland indicator cover by plot is shown in @fig-mon-upl-obl, with
native versus invasive composition in @fig-mon-invasive and noxious weed
occurrence in @fig-mon-noxious.

```{r}
#| label: fig-mon-upl-obl
#| fig-cap: "Wetland indicator cover by plot, current monitoring event."
show_fig("fig_upl_obl_by_plot.png", "wetland indicator cover by plot")
```

```{r}
#| label: fig-mon-invasive
#| fig-cap: "Native vs invasive cover by plot."
show_fig("fig_native_invasive_by_plot.png", "native vs invasive cover by plot")
```

```{r}
#| label: fig-mon-noxious
#| fig-cap: "Noxious weed species heat map."
show_fig("fig_noxious_heatmap.png", "noxious weed heat map")
```

## Hydrology

*[Insert hydrology / water permanence figures and interpretation for the
monitoring period.]*

## Soils

```{r}
#| label: fig-mon-soils
#| fig-cap: "Peat profiles by Von Post humification."
show_fig("fig_peat_vonpost.png", "Von Post peat profiles")
```

# Comparison to Baseline and Performance Targets

*[Compare current-event metrics against baseline (`r params$baseline_year`) and the
program performance targets. A trend figure across monitoring events belongs
here.]*

```{r}
#| label: fig-mon-trend
#| fig-cap: "Vegetation metrics across monitoring events relative to performance targets."
show_fig("fig_monitoring_trend.png", "monitoring trend across events")
```

# Trend Assessment

*[Interpret whether each metric is trending toward or away from target condition,
and whether the trajectory is consistent with monitoring objectives.]*

# Conclusions and Adaptive Management Recommendations

*[State the condition relative to targets, the assessed trend, and any adaptive
management actions recommended for the next monitoring cycle.]*

---

*Report prepared by `r params$prepared_by` for `r params$client`.*
