---
title: "Soils Data Transformation & Classification"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(here)
```

## Load clean data and reference tables

```{r load}
soils <- readRDS(here("data-raw", paste0(params$project_id, "_soils_clean.rds")))
load(here("data", "munsell.rda"))
load(here("data", "von_post.rda"))
load(here("data", "cssc_great_groups.rda"))
```

## Join Munsell hex colour

```{r join-munsell}
soils <- soils |>
  left_join(munsell |> select(hue, value, chroma, hex),
            by = c("munsell_hue" = "hue", "munsell_value" = "value",
                   "munsell_chroma" = "chroma")) |>
  rename(munsell_hex = hex)
```

## Join Von Post humification descriptions

```{r join-vonpost}
soils <- soils |>
  left_join(von_post, by = c("von_post" = "von_post_code"))
```

## Derive material type from horizon code and Von Post score

Material classification (peat / muck / marl / min) was previously a separate
checkbox on the soils field form. It has been moved into derivation here so
the field crew records only what they observe directly — the master horizon
letter from the CSSC code and the Von Post humification score — and the
wetland-relevant material call follows from those by rule:

  - Organic master horizon (O) + Von Post H1–H7  → `peat` (fibric / mesic)
  - Organic master horizon (O) + Von Post H8–H10 → `muck` (humic / well-decomposed)
  - "marl" mentioned anywhere in horizon notes   → `marl` (CaCO3 precipitate;
                                                          not derivable from code)
  - Everything else                              → `min`  (mineral horizon)

If the field crew identifies marl or another non-rule-following material,
they record it in the horizon's notes column and this block picks it up.
The derivation runs before the peat-grade classification so downstream blocks
can rely on `material_type` being populated for every row.

```{r derive-material}
soils <- soils |>
  mutate(
    horizon_master = toupper(stringr::str_extract(horizon_code, "^[A-Za-z]+")),
    material_type  = dplyr::case_when(
      grepl("marl", notes, ignore.case = TRUE)                         ~ "marl",
      startsWith(horizon_master, "O") & von_post %in% paste0("H", 8:10) ~ "muck",
      startsWith(horizon_master, "O")                                   ~ "peat",
      TRUE                                                              ~ "min"
    )
  )
```

## Classify peat horizons — retail vs professional grade

```{r classify-peat}
# Professional peat:        Von Post H1–H4  (fibric to hemic — premium horticultural grade)
# Retail peat:              Von Post H5–H6  (mesic — commercial / bagged garden grade)
# Not commercially viable:  Von Post H7–H10 (heavily decomposed; muck) — peat_grade = NA
# Muck horizons enter peat_thickness accounting (they are still peat material), but
# no commercial grade is assigned because they are not harvestable as product.
# Source: industry convention; refine against project-specific criteria.
soils <- soils |>
  mutate(
    peat_grade = case_when(
      material_type %in% c("peat", "muck") & von_post %in% paste0("H", 1:4) ~ "Professional",
      material_type %in% c("peat", "muck") & von_post %in% paste0("H", 5:6) ~ "Retail",
      TRUE ~ NA_character_
    )
  )
```

## Calculate peat thickness per plot

```{r peat-thickness}
peat_summary <- soils |>
  filter(material_type %in% c("peat", "muck")) |>
  group_by(plot_id, peat_grade) |>
  summarise(thickness_cm = sum(depth_bot_cm - depth_top_cm, na.rm = TRUE),
            .groups = "drop")

peat_summary
```

## Classify plots into CSSC great groups

```{r classify-cssc}
# TODO: implement CSSC key logic using cssc_great_groups reference table
# Requires: presence of organic horizon, depth, saturation indicators
```

## Save

```{r save}
saveRDS(soils,        here("data-raw", paste0(params$project_id, "_soils_transformed.rds")))
saveRDS(peat_summary, here("data-raw", paste0(params$project_id, "_peat_summary.rds")))
```
