---
title: "Water Permanence Data Ingestion"
date: "`r Sys.Date()`"
output: html_document
params:
  project_id: "PROJECT_ID"
  imagery_dir: "data-raw/imagery"
  lidar_dir:   "data-raw/lidar"
  climate_dir: "data-raw/climate"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(terra)
library(sf)
library(dplyr)
library(here)
```

## Notes on data types

This tool handles two structurally different data types:

- **Discrete (snapshot):** Satellite/aerial imagery scenes — each is one moment in time.
  Water presence is classified per-pixel per-scene.
- **Continuous:** Climate data (precipitation, temperature) — time series that must be
  aggregated to match imagery acquisition dates before comparison.

These are kept separate through ingestion and only joined at the analysis stage.

## Imagery — catalogue available scenes

```{r imagery-catalogue}
# List all raster scenes in the imagery directory
img_files <- list.files(here(params$imagery_dir), pattern = "\\.tif$",
                        full.names = TRUE, recursive = TRUE)

img_catalogue <- data.frame(
  path       = img_files,
  filename   = basename(img_files),
  # TODO: parse acquisition date from filename convention (YYYYMMDD expected)
  acq_date   = as.Date(NA)
)

img_catalogue
```

## LiDAR — load DEM / DTM

```{r lidar-load}
# TODO: load and mosaic LiDAR tiles; derive flow accumulation and depression rasters
# dem <- terra::rast(list.files(here(params$lidar_dir), pattern="\\.tif$", full.names=TRUE))
```

## Climate — load precipitation/temperature time series

```{r climate-load}
# TODO: load climate CSV (ECCC station or gridded product) and parse date column
# climate_raw <- read.csv(here(params$climate_dir, "climate.csv"))
```

## Save catalogue

```{r save}
saveRDS(img_catalogue, here("data-raw", paste0(params$project_id, "_imagery_catalogue.rds")))
```
